---
title: "Final Technical Appendix"
editor: visual
---

# Predicting Philadelphia Property Tax Delinquency

## Part 1: Loading and Cleaning Data

```{r setup-libraries}
#| message: false
#| warning: false

install.packages(c(
  "tidyverse",
  "lubridate",
  "sf",
  "tigris",
  "tidycensus",
  "viridis",
  "knitr",
  "kableExtra",
  "patchwork",
  "modelsummary",
  "scales",
  "here",
  "janitor"
))

library(tidyverse)
library(lubridate)
library(sf)
library(tigris)
library(tidycensus)

library(viridis) # color scales for visualization
library(knitr)
library(kableExtra)
library(patchwork)

library(modelsummary)
library(scales)

library(here)
options(scipen = 999) # Get rid of scientific notation to make it look tidier
```

### Property Tax Data

**Clean the tax dataset** 

-   Standardize parcel IDs 

<!-- -->

-   Convert delinquency indicators to binary 

<!-- -->

-   Aggregate by property if multiple years exist 

<!-- -->

-   Create a “most recent status” variable for 2022 

```{r load-data}

prop_tax <- read.csv(here("data/geocoded_output.csv"))

```

```{r clean-data1}

# Columns with mostly NAs
prop_tax %>% 
  count(is.na(unit_type))

prop_tax %>% 
  count(is.na(unit_num))

prop_tax %>% 
  count(is.na(agreement_agency))

prop_tax %>% 
  count(is.na(years_in_bankruptcy))

prop_tax %>% 
  count(is.na(most_recent_bankrupt_year))

prop_tax %>% 
  count(is.na(oldest_bankrupt_year))

prop_tax %>% 
  count(is.na(principal_sum_bankrupt_years))

# Remove majority NA columns
prop_tax <- prop_tax %>% 
  select(-c(unit_type, unit_num, agreement_agency, years_in_bankruptcy, most_recent_bankrupt_year, oldest_bankrupt_year, principal_sum_bankrupt_years))
# Further removing of unnecessary variables
prop_tax <- prop_tax %>% 
  select(-c(owner, co_owner, zip_4, mailing_address, return_mail, coll_agency_num_years, coll_agency_most_recent_year, coll_agency_oldest_year, coll_agency_principal_owed, coll_agency_total_owed, total_amount_bankrupt_years))

prop_tax <- prop_tax %>% 
  select(-c(mailing_zip, the_geom, the_geom_webmercator, cartodb_id, objectid, most_recent_payment_date, total_assessment, building_code, detail_building_description, sequestration_enforcement, bankruptcy))

write.csv(prop_tax, here("data/prop_tax.csv"))
```

```{r}
# EDA Step

# FOR OUR OWN RECORD
# liens_sold variables --> turn into dummy variables, 1 and 0
# narrow time window
# filter building category/buidling description

```

### Housing Characteristics Data

```{r}

opa_prop <- read.csv(here("data/opa_properties_for_FINAL.csv"))
```

## Load Spatial Data

### Philly Limits and Neighborhood

```{r Philadelphia}
# Loading city boundary
PHL_LIMITS <-  st_read(here("data/City_Limits.shp"))

# Loading neighborhood boundaries
phil_hood <- st_read(here("data/philadelphia-neighborhoods.shp"))

#filtering - only need neighborhood names
phil_hood <- phil_hood %>%
  select(LISTNAME, geometry) %>%
  rename(Name = LISTNAME) 

phil_hood <- st_transform(phil_hood, crs=2272)
```

```{r property-spatial}

prop_tax_sf <- st_as_sf(
  prop_tax,
  coords = c("longitude", "latitude"),
  crs = 4326,     # assuming coordinates are lon/lat
  remove = FALSE
)

prop_tax_sf <- st_transform(prop_tax_sf, st_crs(PHL_LIMITS))


class(opa_prop$shape)

bad <- which(!grepl("POINT\\(", opa_prop$shape))
opa_prop$shape[bad][1:10]

sum(opa_prop$shape == "")
sum(is.na(opa_prop$shape))

opa_prop_clean <- opa_prop %>%
  filter(!is.na(shape), shape != "")
opa_prop_clean <- opa_prop_clean %>%
  mutate(shape_clean = gsub("^SRID=2272;\\s*", "", shape))

opa_sf <- st_as_sf(opa_prop_clean, wkt = "shape_clean", crs = 2272)

opa_sf <- st_transform(opa_sf, st_crs(PHL_LIMITS))

```

```{r map-properties}
# ADD TITLES
ggplot() +
  geom_sf(data = PHL_LIMITS, fill = "white", color = "gray60") +
  geom_sf(
    data = prop_tax_sf,
    color = "lightpink",
    size = 0.025,
    alpha = 0.6
  ) +
  geom_sf(
    data = opa_sf,
    color = "black", 
    size = 0.025,
    alpha = 0.6
  ) +
  theme_void()
```

```{r cleaning-delinquent-data}
# check if any duplicate delinquent property records
prop_tax %>%
  count(opa_number, sort = TRUE)


prop_tax <- prop_tax %>%
  # remove entries where op_number = 0
  filter(opa_number != 0) %>% 
  # reduce time window to properties that entered delinquency between 2010 and 2020
  filter(oldest_year_owed >= 2010, oldest_year_owed <= 2020) %>% 
  filter(general_building_description == "house")

# distribution of total amount due for tax delinquent properties
prop_tax %>% 
  ggplot() +
  geom_histogram(aes(x = total_due), binwidth = 1000) +
  coord_cartesian(xlim = c(0, 100000))

# add delinquency indicator to dataframe - all properties assigned 1 because they are all delinquent in this dataset
prop_tax$delinquent <- 1

```

```{r clean-opa}

#filter data to only select Single-Family property types 
  #defined where category_code_description is Single Family
single_prop <- opa_prop_clean %>%
  filter(category_code_description == "SINGLE FAMILY")

#removing NAs and any properties where sales price was equal to or below 0
single_prop_clean <- single_prop %>% 
  filter(!is.na(total_livable_area), !is.na(sale_price), !is.na(year_built), !is.na(location)) %>% 
  filter(sale_price>0)

# check if any duplicate properties
single_prop_clean %>%
  count(parcel_number, sort = TRUE)


#recording IQR of sale price variable for better sense of appropriate thresholds
  #of sale prices that would allow us to discount extreme outlier values
IQR_sale_price <- summary(single_prop_clean$sale_price)

#further narrowing down properties to those with sales prices 
  #between 20,000 and 1,000,000
single_prop_clean <- single_prop_clean %>% 
  filter((sale_price >= 20000 & sale_price <= 1000000))

#histogram of sales price, without outlier values
ggplot(single_prop_clean, aes(x=sale_price)) +
  geom_histogram(fill = "lightsteelblue1", bins = 500) +
  labs(title = "Histogram of Sales Prices", x = "Price", y = "Frequency")

##log transformation of sale price variable data to enhance normality
  #because histogram of un-logged values is right skewed
#p_single_recent <- p_single_recent %>% 
  #mutate(log_sale_price = log(sale_price))

##histogram of log-transformed sale price
#ggplot(p_single_recent, aes(x=log_sale_price)) +
  #geom_histogram(fill = "lightsteelblue1", bins = 200) +
  #labs(title = "Histogram of Log-Transformed Sales Prices", x = "Sale Price", y = "Frequency") +
  #scale_x_continuous(
    #breaks = c(10, 11, 12, 13, 14),
    #labels = function(x) scales::dollar(exp(x))) #change x-axis values to dollars   

#further filtering properties where total area and total liveable area are greater than 0
single_prop_clean <- single_prop_clean %>% 
  filter(total_livable_area > 0)

#labeling empty rows of character variable with NA
house <- single_prop_clean %>% 
  mutate(across(where(is.character), ~ na_if(., "")))

#continued filtering of properties where year built is greater than 0
house <- house %>% 
  filter(year_built > 0)

#saving cleaned single-family home properties data
write_csv(house, here("data/houses_clean.csv"))

#reading cleaned dataframe back into project
house <- read.csv(here("data/houses_clean.csv"))
```

```{r merging-delinquent-to-allproperties}

#### MAKE SURE TO CHANGE COMMENTS IN OUR OWN WORDS

### STEP 1 — Expand delinquency ranges into one row per delinquent year
prop_tax_expanded <- prop_tax %>%
  # Create all combinations of opa_number × year
  expand_grid(year = oldest_year_owed:most_recent_year_owed) %>%
  mutate(delinquent = 1) %>%
  select(opa_number, year, delinquent)

### STEP 2 — Create the full property-year panel (2010–2020)
years <- 2010:2020

panel <- expand_grid(
  house,          # 2024 property data
  year = years
) %>%
  rename(opa_number = parcel_number)

### STEP 3 — Merge expanded delinquency data onto the panel
panel <- panel %>%
  left_join(prop_tax_expanded, by = c("opa_number", "year"))

# checking if merge worked
panel %>% 
  filter(opa_number == "171207000") %>% 
  arrange(year) %>% 
  select(opa_number, year, delinquent)

### STEP 4 — Convert NA delinquency to 0
panel <- panel %>%
  mutate(delinquent = ifelse(is.na(delinquent), 0, 1))

### STEP 5 — Inspect a property to confirm correctness
panel %>%
  filter(opa_number == "171207000") %>%
  arrange(year) %>%
  select(opa_number, year, delinquent)
```

```{r lag-variables}

panel <- panel %>%
  arrange(opa_number, year) %>%
  group_by(opa_number) %>%
  mutate(
    # delinquency in the previous year (0 for first year)
    lag_delinquent = lag(delinquent, default = 0),

    # running count of all past delinquencies
    times_delinquent_before = cumsum(lag_delinquent)
  ) %>%
  ungroup() 

#panel <- panel %>% 
  #select(-ever_delinquent_before)

sum(panel$delinquent == 1)
sum(panel$delinquent == 0)


```

### CENSUS CHARACTERISTICS AT BLOCK GROUP LEVEL

```{r census}

# retrieving census data
#note: the table names and numbers and explanation was suggested by Claude AI, coding domne by team 
pa_county_data <- get_acs(
  geography = "block group",
  state     = "PA",
  county    = "Philadelphia",
  variables = c(
    MEDHHINC    = "B19013_001",  # Median household income
    WHITE       = "B03002_003",  # Non-Hispanic White population
    TOTAL_OCC   = "B25003_001",  # Total occupied housing units
    OWNER_OCC   = "B25003_002",  # Owner occupied units
    RENTER_OCC  = "B25003_003",  # Renter occupied units
    TOTAL_HSG   = "B25001_001",  # Total housing units
    VACANT      = "B25004_001",  # Vacant housing units
    OWN_TOT     = "B25038_002",  # Total owner occupied (move-in universe)
    OWN_IN17P   = "B25038_003",  # Owners moved in 2017 or later
    OWN_IN1516  = "B25038_004",  # Owners moved in 2015-2016
    OWN_IN1014  = "B25038_005",  # Owners moved in 2010-2014
    OWN_IN0009  = "B25038_006",  # Owners moved in 2000-2009
    OWN_IN9099  = "B25038_007",  # Owners moved in 1990-1999
    OWN_INPRE89 = "B25038_008",  # Owners moved in 1989 or earlier
    RNT_TOT     = "B25038_009",  # Total renter occupied (move-in universe)
    RNT_IN17P   = "B25038_010",  # Renters moved in 2017 or later
    RNT_IN1516  = "B25038_011",  # Renters moved in 2015-2016
    RNT_IN1014  = "B25038_012",  # Renters moved in 2010-2014
    RNT_IN0009  = "B25038_013",  # Renters moved in 2000-2009
    RNT_IN9099  = "B25038_014",  # Renters moved in 1990-1999
    RNT_INPRE89 = "B25038_015",  # Renters moved in 1989 or earlier
    BUILT_TOT   = "B25034_001",  # Total housing units (year built universe)
    BUILT_2014P = "B25034_002",  # Built 2014 or later
    BUILT_1013  = "B25034_003",  # Built 2010-2013
    BUILT_0009  = "B25034_004",  # Built 2000-2009
    BUILT_9099  = "B25034_005",  # Built 1990-1999
    BUILT_8089  = "B25034_006",  # Built 1980-1989
    BUILT_7079  = "B25034_007",  # Built 1970-1979
    BUILT_6069  = "B25034_008",  # Built 1960-1969
    BUILT_5059  = "B25034_009",  # Built 1950-1959
    BUILT_4049  = "B25034_010",  # Built 1940-1949
    BUILT_PRE39 = "B25034_011",  # Built 1939 or earlier
    TPOP        = "B01003_001",  # Total population
    M_6566      = "B01001_020",  # Male 65-66
    M_6769      = "B01001_021",  # Male 67-69
    M_7074      = "B01001_022",  # Male 70-74
    M_7579      = "B01001_023",  # Male 75-79
    M_8084      = "B01001_024",  # Male 80-84
    M_85P       = "B01001_025",  # Male 85+
    F_6566      = "B01001_044",  # Female 65-66
    F_6769      = "B01001_045",  # Female 67-69
    F_7074      = "B01001_046",  # Female 70-74
    F_7579      = "B01001_047",  # Female 75-79
    F_8084      = "B01001_048",  # Female 80-84
    F_85P       = "B01001_049"   # Female 85+
  ),
  year    = 2019,
  survey  = "acs5",
  output  = "wide"
)

# calculating margin of error (MOE / Estimate * 100 = coefficient of variation)
pa_county_data <- pa_county_data %>%
  mutate(
    MEDHHINC_MOE   = (MEDHHINCM    / MEDHHINCE)   * 100,  # CV for median household income
    WHITE_MOE      = (WHITEM       / WHITEE)       * 100,  # CV for white population
    TOTAL_OCC_MOE  = (TOTAL_OCCM   / TOTAL_OCCE)  * 100,  # CV for total occupied units
    OWNER_OCC_MOE  = (OWNER_OCCM   / OWNER_OCCE)  * 100,  # CV for owner occupied units
    RENTER_OCC_MOE = (RENTER_OCCM  / RENTER_OCCE) * 100,  # CV for renter occupied units
    TOTAL_HSG_MOE  = (TOTAL_HSGM   / TOTAL_HSGE)  * 100,  # CV for total housing units
    VACANT_MOE     = (VACANTM      / VACANTE)      * 100,  # CV for vacant units
    OWN_TOT_MOE    = (OWN_TOTM     / OWN_TOTE)    * 100,  # CV for owner move-in universe
    RNT_TOT_MOE    = (RNT_TOTM     / RNT_TOTE)    * 100,  # CV for renter move-in universe
    BUILT_TOT_MOE  = (BUILT_TOTM   / BUILT_TOTE)  * 100,  # CV for year built universe
    TPOP_MOE       = (TPOPM        / TPOPE)        * 100   # CV for total population
  )

# categorizing reliability based on CV thresholds
filter_pa_category <- pa_county_data %>%
  mutate(
    reliability = case_when(
      # Unreliable: estimate is zero or missing (CV is NA or infinite)
      is.na(MEDHHINC_MOE)   | is.infinite(MEDHHINC_MOE)   |
      is.na(TOTAL_HSG_MOE)  | is.infinite(TOTAL_HSG_MOE)  |
      is.na(VACANT_MOE)     | is.infinite(VACANT_MOE)     |
      is.na(TPOP_MOE)       | is.infinite(TPOP_MOE)       |
      is.na(OWNER_OCC_MOE)  | is.infinite(OWNER_OCC_MOE)  |
      is.na(RENTER_OCC_MOE) | is.infinite(RENTER_OCC_MOE) ~ "Unreliable",

      # Use with caution: CV exceeds 40% on any core variable
      MEDHHINC_MOE   > 40 |
      TOTAL_HSG_MOE  > 40 |
      VACANT_MOE     > 40 |
      TPOP_MOE       > 40 |
      OWNER_OCC_MOE  > 40 |
      RENTER_OCC_MOE > 40 ~ "Use with Caution",

      TRUE ~ "Reliable"
    )
  ) %>%
  drop_na()

# creating final dataset with derived rates and proportions
final_dataset <- filter_pa_category %>%
  mutate(
    OWNER_RATE      = (OWNER_OCCE  / TOTAL_OCCE)  * 100,  # % of occupied units that are owner occupied
    RENTER_RATE     = (RENTER_OCCE / TOTAL_OCCE)  * 100,  # % of occupied units that are renter occupied
    VACANCY_RATE    = (VACANTE     / TOTAL_HSGE)  * 100,  # % of all units that are vacant
    WHITE_PROP      = (WHITEE      / TPOPE)        * 100,  # % of population that is Non-Hispanic White

    # Avg years owners have lived in unit (weighted midpoint from move-in brackets)
    AVG_OWN_STAY = (
      OWN_IN17PE   * (2019 - 2018) +  # midpoint 2017-2019
      OWN_IN1516E  * (2019 - 2015) +  # midpoint 2015-2016
      OWN_IN1014E  * (2019 - 2012) +  # midpoint 2010-2014
      OWN_IN0009E  * (2019 - 2004) +  # midpoint 2000-2009
      OWN_IN9099E  * (2019 - 1994) +  # midpoint 1990-1999
      OWN_INPRE89E * (2019 - 1984)    # assumed midpoint pre-1989
    ) / OWN_TOTE,

    # Avg years renters have lived in unit (weighted midpoint from move-in brackets)
    AVG_RNT_STAY = (
      RNT_IN17PE   * (2019 - 2018) +  # midpoint 2017-2019
      RNT_IN1516E  * (2019 - 2015) +  # midpoint 2015-2016
      RNT_IN1014E  * (2019 - 2012) +  # midpoint 2010-2014
      RNT_IN0009E  * (2019 - 2004) +  # midpoint 2000-2009
      RNT_IN9099E  * (2019 - 1994) +  # midpoint 1990-1999
      RNT_INPRE89E * (2019 - 1984)    # assumed midpoint pre-1989
    ) / RNT_TOTE,

    # Avg age of housing stock in years (weighted midpoint from year built brackets)
    AVG_HOUSING_AGE = (
      BUILT_2014PE * (2019 - 2016) +  # midpoint 2014-2019
      BUILT_1013E  * (2019 - 2011) +  # midpoint 2010-2013
      BUILT_0009E  * (2019 - 2004) +  # midpoint 2000-2009
      BUILT_9099E  * (2019 - 1994) +  # midpoint 1990-1999
      BUILT_8089E  * (2019 - 1984) +  # midpoint 1980-1989
      BUILT_7079E  * (2019 - 1974) +  # midpoint 1970-1979
      BUILT_6069E  * (2019 - 1964) +  # midpoint 1960-1969
      BUILT_5059E  * (2019 - 1954) +  # midpoint 1950-1959
      BUILT_4049E  * (2019 - 1944) +  # midpoint 1940-1949
      BUILT_PRE39E * (2019 - 1930)    # assumed midpoint pre-1939
    ) / BUILT_TOTE,

    # Sum all male and female 65+ age bands
    POP_65P     = M_6566E + M_6769E + M_7074E + M_7579E + M_8084E + M_85PE +
                  F_6566E + F_6769E + F_7074E + F_7579E + F_8084E + F_85PE,
    OVER65_PROP = (POP_65P / TPOPE) * 100  # % of population aged 65 or older

  ) %>%
  select(
    GEOID, NAME, reliability,
    MEDHHINCE,       # Median household income
    WHITE_PROP,      # % Non-Hispanic White
    OWNER_RATE,      # % owner occupied
    RENTER_RATE,     # % renter occupied
    VACANCY_RATE,    # % vacant units
    AVG_OWN_STAY,    # Avg years owners have stayed
    AVG_RNT_STAY,    # Avg years renters have stayed
    AVG_HOUSING_AGE, # Avg age of housing stock in years
    OVER65_PROP      # % of population 65 or older
  )

head(final_dataset)
```

### Economic Variables from Census AT CENSUS TRACT LEVEL

```{r}

#get census tract level data that we apply to block group level


 #── Census tract-level socioeconomic indicators ──────────────────────────────
# Philadelphia County, ACS 5-year 2019
# Variables: unemployment, educational attainment, poverty,
#            housing cost burden, and SNAP receipt

philly_tract_data <- get_acs(
  geography = "tract",
  state     = "PA",
  county    = "Philadelphia",
  variables = c(
    # ── Labor force / unemployment (B23025) ───────────────────────────────────
    LBR_FORCE    = "B23025_003",  # Civilian labor force
    UNEMPLOYED   = "B23025_005",  # Unemployed

    # ── Educational attainment 25+ (B15003) ───────────────────────────────────
    EDUC_TOT     = "B15003_001",  # Total population 25+
    BACH         = "B15003_022",  # Bachelor's degree
    MASTERS      = "B15003_023",  # Master's degree
    PROF         = "B15003_024",  # Professional school degree
    DOCT         = "B15003_025",  # Doctorate degree

    # ── Poverty (B17001) ──────────────────────────────────────────────────────
    POV_TOT      = "B17001_001",  # Total population for poverty determination
    POV_BELOW    = "B17001_002",  # Income below poverty level

    # ── Housing cost burden — owners (B25091) ─────────────────────────────────
    OWN_BURDEN_TOT = "B25091_001", # Total owner-occupied units w/ mortgage costs
    OWN_30_34    = "B25091_008",  # 30.0–34.9% of income
    OWN_35_39    = "B25091_009",  # 35.0–39.9% of income
    OWN_40_49    = "B25091_010",  # 40.0–49.9% of income
    OWN_50P      = "B25091_011",  # 50.0% or more of income

    # ── Housing cost burden — renters (B25070) ────────────────────────────────
    RNT_BURDEN_TOT = "B25070_001", # Total renter-occupied units w/ gross rent
    RNT_30_34    = "B25070_007",  # 30.0–34.9% of income
    RNT_35_39    = "B25070_008",  # 35.0–39.9% of income
    RNT_40_49    = "B25070_009",  # 40.0–49.9% of income
    RNT_50P      = "B25070_010",  # 50.0% or more of income

    # ── SNAP / food stamp receipt (B22003) ────────────────────────────────────
    SNAP_TOT     = "B22003_001",  # Total households
    SNAP_RCVD    = "B22003_002"   # Households that received SNAP
  ),
  year   = 2019,
  survey = "acs5",
  output = "wide"
)


# ── Coefficient of variation (MOE / Estimate × 100) ──────────────────────────
philly_tract_data <- philly_tract_data %>%
  mutate(
    UNEMP_CV       = (UNEMPLOYEDM   / UNEMPLOYEDE)     * 100,
    EDUC_CV        = (EDUC_TOTM     / EDUC_TOTE)       * 100,
    POV_CV         = (POV_TOTM      / POV_TOTE)        * 100,
    OWN_BURDEN_CV  = (OWN_BURDEN_TOTM / OWN_BURDEN_TOTE) * 100,
    RNT_BURDEN_CV  = (RNT_BURDEN_TOTM / RNT_BURDEN_TOTE) * 100,
    SNAP_CV        = (SNAP_TOTM     / SNAP_TOTE)       * 100
  )


# ── Reliability classification (CV threshold: 40%) ───────────────────────────
philly_tract_filtered <- philly_tract_data %>%
  mutate(
    reliability = case_when(
      # Unreliable: estimate is zero/missing → CV is NA or infinite
      is.na(UNEMP_CV)      | is.infinite(UNEMP_CV)      |
      is.na(EDUC_CV)       | is.infinite(EDUC_CV)       |
      is.na(POV_CV)        | is.infinite(POV_CV)        |
      is.na(OWN_BURDEN_CV) | is.infinite(OWN_BURDEN_CV) |
      is.na(RNT_BURDEN_CV) | is.infinite(RNT_BURDEN_CV) |
      is.na(SNAP_CV)       | is.infinite(SNAP_CV)
      ~ "Unreliable",

      # Use with caution: any core denominator CV exceeds 40%
      UNEMP_CV      > 40 |
      EDUC_CV       > 40 |
      POV_CV        > 40 |
      OWN_BURDEN_CV > 40 |
      RNT_BURDEN_CV > 40 |
      SNAP_CV       > 40
      ~ "Use with Caution",

      TRUE ~ "Reliable"
    )
  ) %>%
  drop_na()


# ── Derived rates ─────────────────────────────────────────────────────────────
final_tract_dataset <- philly_tract_filtered %>%
  mutate(
    # Unemployment rate: unemployed / civilian labor force
    UNEMP_RATE     = (UNEMPLOYEDE  / LBR_FORCEE)       * 100,

    # Bachelor's degree or higher: sum of bach + grad degrees / pop 25+
    BACH_PLUS      = BACHE + MASTERSE + PROFE + DOCTE,
    BACH_RATE      = (BACH_PLUS    / EDUC_TOTE)        * 100,

    # Poverty rate: population below poverty / total poverty universe
    POV_RATE       = (POV_BELOWE   / POV_TOTE)         * 100,

    # Cost-burdened owners (≥30% income on housing costs)
    OWN_BURDENED   = OWN_30_34E + OWN_35_39E + OWN_40_49E + OWN_50PE,
    OWN_BURDEN_RATE= (OWN_BURDENED / OWN_BURDEN_TOTE)  * 100,

    # Cost-burdened renters (≥30% income on gross rent)
    RNT_BURDENED   = RNT_30_34E + RNT_35_39E + RNT_40_49E + RNT_50PE,
    RNT_BURDEN_RATE= (RNT_BURDENED / RNT_BURDEN_TOTE)  * 100,

    # Overall cost-burden rate (owners + renters combined)
    TOTAL_BURDENED = OWN_BURDENED + RNT_BURDENED,
    TOTAL_BURDEN_DENOM = OWN_BURDEN_TOTE + RNT_BURDEN_TOTE,
    BURDEN_RATE    = (TOTAL_BURDENED / TOTAL_BURDEN_DENOM) * 100,

    # SNAP receipt rate: households receiving SNAP / total households
    SNAP_RATE      = (SNAP_RCVDE   / SNAP_TOTE)        * 100
  ) %>%
  select(
    GEOID, NAME, reliability,
    UNEMP_RATE,       # % unemployed (of civilian labor force)
    BACH_RATE,        # % with bachelor's degree or higher (pop 25+)
    POV_RATE,         # % below poverty line
    OWN_BURDEN_RATE,  # % cost-burdened owner-occupied units
    RNT_BURDEN_RATE,  # % cost-burdened renter-occupied units
    BURDEN_RATE,      # % cost-burdened all occupied units (combined)
    SNAP_RATE         # % households receiving SNAP
  )

head(final_tract_dataset)

```

### Interaction Term: Republican Gov.

```{r}
#creating a variable that is republican or not, then we can multiply this later to see effects

panel_census_geo <- panel_census_geo %>%
  mutate(republican_governor = if_else(year %in% 2011:2014, 1, 0))
         
```

### Business Improvement Districts

```{r BID}
#Business Improvement Districts

bid <- st_read("https://services.arcgis.com/fLeGjb7u4uXqeF9q/arcgis/rest/services/business_improvement_districts/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson")%>%
  st_transform(crs=2272)

bid <- bid %>%
  select(organization, geometry)

#creating centroid for spatial engineering later

#bid_centroid <- st_centroid(bid)


#City of Philadelphia. (n.d.). Business improvement districts [Data set]. OpenDataPhilly.
```

### Affordable Housing Projects

```{r Aff Housing}

#Point Data

#Affordable Housing Projects
aff_house <- st_read("https://hub.arcgis.com/api/v3/datasets/ca8944190b604b2aae7eff17c8dd9ef5_0/downloads/data?format=geojson&spatialRefId=4326&where=1%3D1")%>%
  st_transform(crs= 2272)

aff_house <- aff_house %>%
  filter(status !="Under Construction")%>%
  select(project_type, geometry, developer_name, development_type)

aff_house <- aff_house%>%
  rename(Aff_project = project_type, aff_developer = developer_name, aff_development= development_type)

#buffer is in ft atm CRS= 2272. .5 mile= 2640ft

aff_buf <- st_buffer(aff_house, dis=2640)


#City of Philadelphia. (n.d.). Completed affordable housing units [Data set]. OpenDataPhilly. https://opendataphilly.org/datasets/affordable-housing-production/
```

### Empowerment Zone

```{r Emp Zone}

#Polygon data. 

#Decided not to use centroid because we don't know how large these spaces are and we want to see if the closer you are the more or less likely you are tax delinquent. 

emp_zn <- st_read("https://hub.arcgis.com/api/v3/datasets/78fa563815654742909159b53fa4b065_0/downloads/data?format=geojson&spatialRefId=4326&where=1%3D1")%>%
  st_transform(crs= 2272)

emp_zn <- emp_zn%>%
  select(ZONE_NAME, geometry)%>%
  rename(Emp_Zone = ZONE_NAME)


#City of Philadelphia. (n.d.). Empowerment zones [Data set]. OpenDataPhilly. https://opendataphilly.org/datasets/empowerment-zones/

```

### Housing Counselling Agencies

```{r Housing Councelling Agencies HCA}

#POINT DATA

HCA <- st_read("https://hub.arcgis.com/api/v3/datasets/3265538198254e9fb6a8974745adab51_0/downloads/data?format=geojson&spatialRefId=4326&where=1%3D1")%>%
               st_transform(crs=2272)

HCA<-HCA%>%
  select(AGENCY, geometry)%>%
  rename(HCA_AGENCY = AGENCY)



#City of Philadelphia. (n.d.). Housing counseling agencies [Data set]. OpenDataPhilly. https://opendataphilly.org/datasets/housing-counseling-agencies/
```

### Load and Clean Crime Data

```{r}
library(janitor)
options(tigris_use_cache = TRUE)

# ---- 2. READ THE FILES ----
crime19 <- read_csv(here("data/incidents_part1_part2_2019.csv"), show_col_types = FALSE) %>%
  clean_names()

#crime24 <- read_csv("~/Downloads/incidents_part1_part2.csv", show_col_types = FALSE) %>%
  #clean_names()

# ---- 3. KEEP ONLY THE COLUMNS WE NEED ----
crime19 <- crime19 %>%
  select(dispatch_date, text_general_code, lat, lng)

#crime24 <- crime24 %>%
  #select(dispatch_date, text_general_code, lat, lng)

# ---- 4. PARSE DATES + FILTER TO THE CORRECT YEAR ----
crime19 <- crime19 %>%
  mutate(dispatch_date = as.Date(dispatch_date)) %>%
  filter(format(dispatch_date, "%Y") == "2019")

#crime24 <- crime24 %>%
  #mutate(dispatch_date = as.Date(dispatch_date)) %>%
  #filter(format(dispatch_date, "%Y") == "2024")

# ---- 5. FILTER TO ONE CRIME TYPE ----
# Recommended: combine residential + non-residential burglary
burglary_codes <- c("Burglary Residential", "Burglary Non-Residential")

crime19_burg <- crime19 %>%
  filter(text_general_code %in% burglary_codes) %>%
  filter(!is.na(lat), !is.na(lng))

#crime24_burg <- crime24 %>%
  #filter(text_general_code %in% burglary_codes) %>%
  #filter(!is.na(lat), !is.na(lng))

# ---- 6. CONVERT INCIDENTS TO SF POINTS ----
crime19_sf <- st_as_sf(crime19_burg, coords = c("lng", "lat"), crs = 4326, remove = FALSE)
#crime24_sf <- st_as_sf(crime24_burg, coords = c("lng", "lat"), crs = 4326, remove = FALSE)`
```

### Free Meal Sites Cleaning

```{r freemeal-cleaning}
#| message: false
#| warning: false

#selecting freemeal sites
freemeal <- st_read("https://hub.arcgis.com/api/v3/datasets/5825a32bb8844bb097f7a16d4fbf4f23_0/downloads/data?format=geojson&spatialRefId=4326&where=1%3D1") %>%
  st_transform(2272)

#cleaning and slecting useful columns and removing duplicates
freemeal_clean <- freemeal %>%
  select(
    FM_SITE = site_name,
    FM_ADDRESS = address,
  ) %>%
  distinct(FM_SITE, .keep_all = TRUE)
```

```{r freemeal-engineering}
#| message: false
#| warning: false

#buffers of 5 mile around free meal sites
freemeal_engineered <- freemeal_clean %>%
  mutate(fm_buffer_1mile = st_buffer(geometry, 26400)
  )
```

```{r summary2}
#| message: false
#| warning: false

#summary table of cleaning process of free meal sites
data.frame(
  Phase = c("Unclean", "Cleaned", "With Buffers"),
  Columns = c(ncol(freemeal), ncol(freemeal_clean), ncol(freemeal_engineered)),
  Rows = c(nrow(freemeal), nrow(freemeal_clean), nrow(freemeal_engineered))
) %>%
  kable(caption = "Summary of Free Meal Sites Data Cleaning") %>%
  kable_styling(
    full_width = FALSE,
    position = "left",
    bootstrap_options = c("striped", "hover", "condensed", "sm")
  )
```

### Parks Cleaning

```{r parkscleaning}
#| message: false
#| warning: false

#loading parks
parks1 <- st_read("https://hub.arcgis.com/api/v3/datasets/d52445160ab14380a673e5849203eb64_0/downloads/data?format=geojson&spatialRefId=4326&where=1%3D1") %>%
  st_transform(crs=2272) 

#filtering all parks classified as "TRAFFIC_ISLAND_MEDIAN" and checking total number of rows
parks1 %>% 
  filter(property_classification == "TRAFFIC_ISLAND_MEDIAN") %>%
  nrow()

# loading and cleaning park data, removing inaccessible classifications
parks <- parks1 %>% 
  filter(!(property_classification == "TRAFFIC_ISLAND_MEDIAN" & ppr_use == "OTHER")) %>%
  filter(!(property_classification == "PATHWAY")) %>% 
  filter(!(property_classification == "OPERATIONAL_INTERNAL")) %>%
  rename(park_area = Shape__Area,
         park_length = Shape__Length) %>%
  select(label, nested, address_911, acreage, property_classification, ppr_use, park_area, park_length, geometry) %>% 
  st_transform(2272)

```

```{r}
#| message: false
#| warning: false

#table of parks dataframe rows and columns throughout cleaning phases
data.frame(
  Phase= c("Unclean", "Final Data"), 
  Column= c(ncol(parks1), ncol(parks)), 
  Rows= c(nrow(parks1), nrow(parks)
))%>%
  kable(caption= "Summary of Cleaned Park Data")%>% 
   kable_styling(
    full_width = FALSE, 
    position = "left",
    bootstrap_options = c("striped", "hover", "condensed", "sm")
  )
```

## Part 2: Spatial Engineering

### Joining Census Data to Property Panel

```{r censusdatajoining}

#calling block group shape
block_group_shape <- st_read("https://hub.arcgis.com/api/v3/datasets/2f982bada233478ea0100528227febce_0/downloads/data?format=geojson&spatialRefId=4326&where=1%3D1")%>%
  st_transform(crs = 2272)

census_data <- final_dataset

block_group_shape <- block_group_shape %>%
  rename(GEOID = GEOID10)

block_group_shape <- block_group_shape %>%
  mutate(GEOID = as.character(GEOID)) 
census_data <- census_data %>% 
  mutate(GEOID = as.character(GEOID))


census_with_geography <- block_group_shape %>% 
  left_join(census_data, by= "GEOID")

census_with_geography <- st_transform (census_with_geography, 2272)

panel_sf <- st_as_sf(panel, 
                     wkt = "shape_clean", crs = 2272)
  

panel_census_geo <- st_join(panel_sf, census_with_geography, join = st_within)

```

```{r censusdatajopining}

#used claude to attach census tract level data to block group level in the dataset. we wanted to ensure that economic data (absent from block group level) was still included in our prediction for accuracy. 

census_tract_level <- st_read("https://hub.arcgis.com/api/v3/datasets/8bc0786524a4486bb3cf0f9862ad0fbf_0/downloads/data?format=geojson&spatialRefId=4326&where=1%3D1") %>%
  st_transform(crs = 2272) %>%
  rename(GEOID = GEOID10) %>%
  mutate(GEOID = as.character(GEOID))

econ_data <- final_tract_dataset %>%
  mutate(GEOID = as.character(GEOID))

econ_with_geography <- census_tract_level %>%
  left_join(econ_data, by = "GEOID")

block_group_with_econ <- block_group_shape %>%
  st_centroid() %>%                              # use centroids to avoid boundary issues
  st_join(econ_with_geography %>% select(-GEOID), join = st_within) %>%  # spatial join to tracts
  st_drop_geometry() %>%                         # drop centroid geometry
  left_join(block_group_shape, by = "GEOID") %>% # re-attach original block group polygons
  st_as_sf()

panel_census_geo <- st_join(panel_sf, block_group_with_econ, join = st_within)
```

\

### Are any of the houses within Business Improvement Districts??????

```{r bidjoining }

panel_census_geo <- st_transform(panel_census_geo, 2272)
bid <- st_transform(bid, 2272)

panel_census_geo$in_bid <- as.integer(lengths(st_within(panel_census_geo, bid)) > 0)

```

### Properties within 1mile of Affordable Housing Sites

```{r}

panel_census_geo$in_aff_buf <- as.integer(lengths(st_within(panel_census_geo, aff_buf)) > 0)
```

### Homes within Empowerment

```{r}
panel_census_geo$in_emp <- as.integer(lengths(st_within(panel_census_geo, emp_zn)) > 0)
```

### Housing Counselling Agency

```{r}

#distance in ft

panel_census_geo$dist_HCA <- as.numeric(st_distance(panel_census_geo, st_union(HCA)))
```

### Free Meal Sites -distance

```{r}

#distance to mile

panel_census_geo$dist_fm_buf <- as.numeric(st_distance(panel_census_geo, st_union(freemeal_clean)))
```

### Distance to Parks

```{r}


#distance in ft

panel_census_geo$dist_parks <- as.numeric(st_distance(panel_census_geo, st_union(parks)))
```

### Neighborhood Fixed Effects

```{r}
panel_census_geo <- st_transform(panel_census_geo, 2272)
phil_hood <- st_transform(phil_hood, 2272)

panel_census_geo <- st_join(panel_census_geo, phil_hood, join = st_within)
```

```{r}


final_data_panel <- panel_census_geo %>%
  select(
    opa_number, GEOID, category_code_description, location, mailing_city_state, 
    census_tract, NAMELSAD10.x, market_value, sale_price, year, year_built, 
    zoning, shape_clean, total_livable_area, exterior_condition, delinquent, 
    lag_delinquent, times_delinquent_before, republican_governor, UNEMP_RATE, 
    BACH_RATE, POV_RATE, OWN_BURDEN_RATE, RNT_BURDEN_RATE, SNAP_RATE, in_bid, 
    in_aff_buf, in_emp, dist_HCA, dist_parks, dist_fm_buf, Name
  )



```

## Part 3: Exploratory Data Analysis

# 
