---
title: "Bikeshare Space-Time Prediction"
subtitle: "CPLN 5080 - Spring 2026"
author: "Your Name"
date: today
format:
  html:
    code-fold: show
    code-tools: true
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

# INTRODUCTION

## The Rebalancing Challenge in Philadelphia

Philadelphia's Indego bike share system faces the same operational challenge as every bike share system: rebalancing bikes to meet anticipated demand.

At 6:00 AM on a Monday morning, an Indego operations manager could be in charge of managing 200 stations across Philadelphia. With limited trucks and staff for moving bikes 2-3 hours before morning rush hour demand peaks, it is crucial to have a confident idea of which stations will run out of bikes by 8:30 AM.

This lab will build predictive models that forecast bike share demand across space (different stations) and time (different hours) to help solve this operational problem.

## Learning Objectives

1.  **Understand panel data structure** for space-time analysis
2.  **Create temporal lag variables** to capture demand persistence
3.  **Build multiple predictive models** with increasing complexity
4.  **Validate models temporally** (train on past, test on future)
5.  **Analyze prediction errors** in both space and time
6.  **Engineer new features** based on error patterns
7.  **Critically evaluate** when prediction errors matter most

```{r setup}
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(sf)
library(tigris)
library(tidycensus)
library(riem)  # For Philadelphia weather from ASOS stations
library(suncalc) # For calculating daylight hours (sunset and sunrise)

library(viridis) # color scales for visualization
library(knitr)
library(kableExtra)
library(patchwork)


library(here)
options(scipen = 999) # Get rid of scientific notation to make it look tidier

```

### Set up

```{r themes}
plotTheme <- theme_minimal() +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption  = element_text(size = 8),
    axis.text.x   = element_text(size = 10, angle = 45, hjust = 1),
    axis.text.y   = element_text(size = 10),
    axis.title    = element_text(size = 11, face = "bold"),
    panel.grid.major = element_line(colour = "#D0D0D0", linewidth = 0.2),
    panel.grid.minor = element_blank(),
    axis.ticks    = element_blank(),
    legend.position = "right"
  )

mapTheme <- theme_void() +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption  = element_text(size = 8),
    legend.position  = "right",
    plot.margin      = margin(1, 1, 1, 1, "cm"),
    legend.key.height = unit(1, "cm"),
    legend.key.width  = unit(0.2, "cm")
  )

palette5 <- c("#eff3ff", "#bdd7e7", "#6baed6", "#3182bd", "#08519c")
```

# IMPORT AND PREP DATA

## Load Indego Trip Data (Q1 2025)

```{r load_indego}
#| message: false
#| warning: false
#| results: hide

# Read Q1 2025 data
indego <- read_csv(here("data/indego-trips-2025-q1.csv"))

# Quick look at the data
glimpse(indego)
```

After a quick glance, this raw dataframe from the first quarter of 2025 shows records of each bike trip that occurred during the specified timeline. The geographic locations of each start station are given by the columns, 'start_lat' and 'start_lon' whose values look to be in degree units. This indicates that the data is currently in the WGS 84 (World Geodetic System 1984) geographic coordinate system, which represents locations as latitude and longitude in degrees. Rather than separate columns, the date and time of each trip are recorded together in both the columns, 'start_time' and 'end_time'.

## Examine the Data Structure

```{r explore_data}
#| message: false
#| warning: false

# How many trips?
cat("Total trips in Q1 2025:", nrow(indego), "\n")

# Date range - Unix timestamp (seconds since 1970)
cat("Date range:", 
    min(mdy_hm(indego$start_time)), "to", 
    max(mdy_hm(indego$start_time)), "\n")

# How many unique stations?
cat("Unique start stations:", length(unique(indego$start_station)), "\n")

# Trip types
table(indego$trip_route_category)

# Passholder types
table(indego$passholder_type)

# Bike types
table(indego$bike_type)
```

## Create Time Bins

To transform the existing dataframe to fit a panel data structure, the trips must be aggregated into hourly intervals. Panel data provides the basis for building a model that can predict the total expected trips per hour per station. The first step towards this final structure is to create time bins using the lubridate library to extract the current date/time data.

```{r create-time-bins}
indego <- indego %>%
  mutate(
    # Parse datetime
    start_datetime = mdy_hm(start_time),
    end_datetime = mdy_hm(end_time),
    
    # Create hourly bins
    interval60 = floor_date(start_datetime, unit = "hour"),
    
    # Extract time features
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE), # day of the week
    hour = hour(interval60),
    date = as.Date(interval60),
    
    # Create useful indicators
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

# Look at temporal features
head(indego %>% select(start_datetime, interval60, week, dotw, hour, weekend))
```

## Trips Over Time

```{r trips_over_time}
#| message: false
#| warning: false

# Daily trip counts
daily_trips <- indego %>%
  group_by(date) %>%
  summarize(trips = n())

ggplot(daily_trips, aes(x = date, y = trips)) +
  geom_line(color = "#3182bd", linewidth = 1) +
  geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
  labs(
    title = "Indego Daily Ridership - Q1 2025",
    subtitle = "Winter demand patterns in Philadelphia",
    x = "Date",
    y = "Daily Trips",
    caption = "Source: Indego bike share"
  ) +
  plotTheme
```

**Question:** What patterns do you see? How does ridership change over time?

Exploring the bike trip data further, we notice a big spike in the number of trips in February. This spike likely aligns with the festivities associated with the Eagles Super Bowl Championship celebration across the city which occurred on Friday, the 14th of February.

We can investigate this further by comparing the 14th to all other Friday's in the data. Indeed, the total Indego bike trips on that day was unusual compared to most other Fridays in the data. This points to the significance of events or holidays for Indego bike usage - something to take into account when deciding which variables may be most useful for predicting demand.

```{r}
fly_eagles_fly <- daily_trips %>% filter(date == "2025-02-14")

typical_boring_friday <- indego %>%
  filter(dotw == "Fri", date != "2025-02-14") %>%
  group_by(date) %>%
  summarize(trips = n()) %>%
  summarize(avg_friday_trips = mean(trips))

print(fly_eagles_fly)
print(typical_boring_friday)

```

## Hourly Patterns

```{r hourly_patterns}
# Average trips by hour and day type
hourly_patterns <- indego %>%
  group_by(hour, weekend) %>%
  summarize(avg_trips = n() / n_distinct(date)) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(hourly_patterns, aes(x = hour, y = avg_trips, color = day_type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Average Hourly Ridership Patterns",
    subtitle = "Clear commute patterns on weekdays",
    x = "Hour of Day",
    y = "Average Trips per Hour",
    color = "Day Type"
  ) +
  plotTheme
```

Generally, the data demonstrate that during weekdays, there are peaks of bike usage which line up with typical commuting hours in the morning and late afternoon. These peaks are dampened during the weekends, but still show a mirroring trend where the number of trips taken increases throughout the daytime and gradually decreases towards the evening hours.

## Top Stations

```{r top_stations}
# Most popular origin stations
top_stations <- indego %>%
  count(start_station, start_lat, start_lon, name = "trips") %>%
  arrange(desc(trips)) %>%
  head(20)

kable(top_stations, 
      caption = "Top 20 Indego Stations by Trip Origins",
      format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# PUT DATA IN CONTEXT

## Load Philadelphia Census Data

We'll get census tract data to add demographic context to our stations.

```{r load_census}
#| message: false
#| warning: false
#| results: hide

# Get Philadelphia census tracts
philly_census <- get_acs(
  geography = "tract",
  variables = c(
    "B01003_001",  # Total population
    "B19013_001",  # Median household income
    "B08301_001",  # Total commuters
    "B08301_010",  # Commute by transit
    "B02001_002",  # White alone
    "B25077_001"   # Median home value
  ),
  state = "PA",
  county = "Philadelphia",
  year = 2022,
  geometry = TRUE,
  output = "wide"
) %>%
  rename(
    Total_Pop = B01003_001E,
    Med_Inc = B19013_001E,
    Total_Commuters = B08301_001E,
    Transit_Commuters = B08301_010E,
    White_Pop = B02001_002E,
    Med_Home_Value = B25077_001E
  ) %>%
  mutate(
    Percent_Taking_Transit = (Transit_Commuters / Total_Commuters) * 100,
    Percent_White = (White_Pop / Total_Pop) * 100
  ) %>%
  st_transform(crs = 4326)  #transforming to WGS84 for lat/lon matching, currently in NAD83

# Check the data
glimpse(philly_census)
```

## Map Philadelphia Context

```{r map_philly}
# Map median income
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = NA) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Context for understanding bike share demand patterns"
  ) +
  # Stations 
  geom_point(
    data = indego,
    aes(x = start_lon, y = start_lat),
    color = "red", size = 0.25, alpha = 0.6
  ) +
  mapTheme
```

Currently, the data are in WGS84. To calculate a spatial distance-based variable, like a buffer that data should be converted to the geographically appropriate projected coordinate system. For Philadelphia, this system is the NAD83(2011) / Pennsylvania South (ftUS). This is a more appropriate coordinate system to calculate accurate measures of distance.

## Join Census Data to Stations

We'll spatially join census characteristics to each bike station.

```{r join_census_to_stations}
# Create sf object for stations
stations_sf <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326)

# Spatial join to get census tract for each station
stations_census <- st_join(stations_sf, philly_census, left = TRUE) %>%
  st_drop_geometry()

# Look at the result - investigate whether all of the stations joined to census data -- according to the map above there are stations in non-residential tracts.

stations_for_map <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Add back to trip data
indego_census <- indego %>%
  left_join(
    stations_census %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )


# Prepare data for visualization
stations_for_map <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Create the map showing problem stations
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = "white", size = 0.1) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar,
    na.value = "grey90"
  ) +
  # Stations with census data (small grey dots)
  geom_point(
    data = stations_for_map %>% filter(has_census),
    aes(x = start_lon, y = start_lat),
    color = "grey30", size = 1, alpha = 0.6
  ) +
  # Stations WITHOUT census data (red X marks the spot)
  geom_point(
    data = stations_for_map %>% filter(!has_census),
    aes(x = start_lon, y = start_lat),
    color = "red", size = 1, shape = 4, stroke = 1.5
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Indego stations shown (RED = no census data match)",
    caption = "Red X marks indicate stations that didn't join to census tracts"
  ) +
  mapTheme



```

## Dealing with missing data

The above map reveals that our dataset contains station locations in non-residential areas. For the sake of simplicity in this lab, we are going to remove them, narrowing our scope to only stations in residential neighborhoods.

```{r}
# Identify which stations to keep
valid_stations <- stations_census %>%
  filter(!is.na(Med_Inc)) %>%
  pull(start_station)

# Filter trip data to valid stations only
indego_census <- indego %>%
  filter(start_station %in% valid_stations) %>%
  left_join(
    stations_census %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )

```

## Get Weather Data

Weather is another variable that may significantly affect bike share demand.

```{r get_weather}
# Get weather from Philadelphia International Airport (KPHL)
# This covers Q1 2025: January 1 - March 31
weather_data <- riem_measures(
  station = "PHL",  # Philadelphia International Airport
  date_start = "2025-01-01",
  date_end = "2025-03-31"
)

# Process weather data
weather_processed <- weather_data %>%
  mutate(
    interval60 = floor_date(valid, unit = "hour"),
    Temperature = tmpf,  # Temperature in Fahrenheit
    Precipitation = ifelse(is.na(p01i), 0, p01i),  # Hourly precip in inches
    Wind_Speed = sknt  # Wind speed in knots
  ) %>%
  select(interval60, Temperature, Precipitation, Wind_Speed) %>%
  distinct(interval60, .keep_all = TRUE)

# Check for missing hours and interpolate if needed
weather_complete <- weather_processed %>%
  # finding gaps: generates every single hour between Jan 1 and Mar 31, like a complete checklist, then compares that checklist against what is actually in weather_processed
  # any hour that has missing data gets added as a new row and data is NA
  complete(interval60 = seq(min(interval60), max(interval60), by = "hour")) %>% 
  # filling the gaps: .direction = "down" means it looks upward to the most recent non-NA value and copies it down into the gap (if 3am is missing, borrows from 2am)
  fill(Temperature, Precipitation, Wind_Speed, .direction = "down") 

# Look at the weather
summary(weather_complete %>% select(Temperature, Precipitation, Wind_Speed))
```

## Visualize Weather Patterns

```{r visualize-weather}
#| message: false
#| warning: false

ggplot(weather_complete, aes(x = interval60, y = Temperature)) +
  geom_line(color = "#3182bd", alpha = 0.7) +
  # se = FALSE: removes confidence interval band around the trend line, default for 
  # geom_smooth()
  geom_smooth(se = FALSE, color = "red") + 
  labs(
    title = "Philadelphia Temperature - Q1 2025",
    subtitle = "Winter to early spring transition",
    x = "Date",
    y = "Temperature (°F)"
  ) +
  plotTheme
```

# CREATE SPACE-TIME PANEL

## Aggregate Trips to Station-Hour Level

```{r aggregate-trips}
#| message: false
#| warning: false

# Count trips by station-hour
trips_panel <- indego_census %>%
  group_by(interval60, start_station, start_lat, start_lon,
           Med_Inc, Percent_Taking_Transit, Percent_White, Total_Pop) %>%
  summarize(Trip_Count = n()) %>%
  ungroup()

# How many station-hour observations?
nrow(trips_panel)

# How many unique stations?
length(unique(trips_panel$start_station))

# How many unique hours?
length(unique(trips_panel$interval60))

hist(trips_panel$Trip_Count)
```

## Create Complete Panel Structure

Not every station has trips every hour. Though, we need a complete panel where every station-hour combination exists (even if Trip_Count = 0). This is accomplished using the expand_grid() function.

```{r complete-panel}
# Calculate expected panel size
n_stations <- length(unique(trips_panel$start_station))
n_hours <- length(unique(trips_panel$interval60))
expected_rows <- n_stations * n_hours

cat("Expected panel rows:", format(expected_rows, big.mark = ","), "\n")
cat("Current rows:", format(nrow(trips_panel), big.mark = ","), "\n")
cat("Missing rows:", format(expected_rows - nrow(trips_panel), big.mark = ","), "\n")

# Create complete panel
study_panel <- expand_grid(
  interval60 = unique(trips_panel$interval60),
  start_station = unique(trips_panel$start_station)
) %>%
  # Join trip counts
  left_join(trips_panel %>% 
              select(interval60, start_station, Trip_Count),
            by = c("interval60", "start_station")) %>%
  # Replace NA trip counts with 0
  mutate(Trip_Count = replace_na(Trip_Count, 0))

glimpse(study_panel)
hist(study_panel$Trip_Count)

# Fill in station attributes (they're the same for all hours)
station_attributes <- trips_panel %>%
  group_by(start_station) %>%
  summarize(
    start_lat = first(start_lat),
    start_lon = first(start_lon),
    Med_Inc = first(Med_Inc),
    Percent_Taking_Transit = first(Percent_Taking_Transit),
    Percent_White = first(Percent_White),
    Total_Pop = first(Total_Pop)
  )

glimpse(station_attributes)

study_panel <- study_panel %>%
  left_join(station_attributes, by = "start_station")

glimpse(study_panel)

# Verify we have complete panel
cat("Complete panel rows:", format(nrow(study_panel), big.mark = ","), "\n")
```

## Add Time Features

```{r add_time_features}
study_panel <- study_panel %>%
  mutate(
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

glimpse(study_panel)
```

## Join Weather Data

```{r join_weather}
study_panel <- study_panel %>%
  left_join(weather_complete, by = "interval60")

glimpse(study_panel)

# Check for missing values
summary(study_panel %>% select(Trip_Count, Temperature, Precipitation))

# there are 5880 NA's for Temperature and Precipitation
```

# CREATE TEMPORAL LAG VARIABLES

The key innovation for space-time prediction: **past demand predicts future demand**.

## Why Temporal Lags?

If there were 15 bike trips from Station A at 8:00 AM, there will probably be \~15 trips at 9:00 AM. We can use this temporal persistence to improve predictions.

**Note:** We still can't use spatial lags in prediction. Temporal lags are different because time moves in one direction - past predicts the future, but the future does not predict the past. Spatial lags are non-directional and create circular prediction logic.

```{r create_lags}
# Sort by station and time
study_panel <- study_panel %>%
  arrange(start_station, interval60)

# Create lag variables WITHIN each station
study_panel <- study_panel %>%
  group_by(start_station) %>%
  mutate(
    lag1Hour = lag(Trip_Count, 1),
    lag2Hours = lag(Trip_Count, 2),
    lag3Hours = lag(Trip_Count, 3),
    lag12Hours = lag(Trip_Count, 12),
    lag1day = lag(Trip_Count, 24)
  ) %>%
  ungroup()

# Remove rows with NA lags (first 24 hours for each station)
study_panel_complete <- study_panel %>%
  filter(!is.na(lag1day))

cat("Rows after removing NA lags:", format(nrow(study_panel_complete), big.mark = ","), "\n")
```

## Visualize Lag Correlations

```{r lag_correlations}
# Sample one station to visualize
example_station <- study_panel_complete %>%
  filter(start_station == first(start_station)) %>%
  head(168)  # One week

# Plot actual vs lagged demand
ggplot(example_station, aes(x = interval60)) +
  geom_line(aes(y = Trip_Count, color = "Current"), linewidth = 1) +
  geom_line(aes(y = lag1Hour, color = "1 Hour Ago"), linewidth = 1, alpha = 0.7) +
  geom_line(aes(y = lag1day, color = "24 Hours Ago"), linewidth = 1, alpha = 0.7) +
  scale_color_manual(values = c(
    "Current" = "#08519c",
    "1 Hour Ago" = "#3182bd",
    "24 Hours Ago" = "#6baed6"
  )) +
  labs(
    title = "Temporal Lag Patterns at One Station",
    subtitle = "Past demand predicts future demand",
    x = "Date-Time",
    y = "Trip Count",
    color = "Time Period"
  ) +
  plotTheme
```

# TEMPORAL TRAIN/TEST SPLIT

## Why Temporal Validation Matters

In real operations, at 6:00 AM on March 15, we need to predict demand for March 15-31. We have data from Jan 1 - March 14, but NOT from March 15-31 (it hasn't happened yet!).

**Wrong approach:** Train on weeks 10-13, test on weeks 1-9 (predicting past from future!)

**Correct approach:** Train on weeks 1-9, test on weeks 10-13 (predicting future from past)

```{r temporal-split}
# Split by week
# Q1 has weeks 1-13 (Jan-Mar)
# Train on weeks 1-9 (Jan 1 - early March)
# Test on weeks 10-13 (rest of March)

# Which stations have trips in BOTH early and late periods?
early_stations <- study_panel_complete %>%
  filter(week < 10) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

late_stations <- study_panel_complete %>%
  filter(week >= 10) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

# Keep only stations that appear in BOTH periods
common_stations <- intersect(early_stations, late_stations)


# Filter panel to only common stations
study_panel_complete <- study_panel_complete %>%
  filter(start_station %in% common_stations)

# NOW create train/test split
train <- study_panel_complete %>%
  filter(week < 10)

test <- study_panel_complete %>%
  filter(week >= 10)

cat("Training observations:", format(nrow(train), big.mark = ","), "\n")
cat("Testing observations:", format(nrow(test), big.mark = ","), "\n")
cat("Training date range:", min(train$date), "to", max(train$date), "\n")
cat("Testing date range:", min(test$date), "to", max(test$date), "\n")

```

## Model 1: Baseline (Time + Weather)

```{r model1}

# Create day of week factor with treatment (dummy) coding
train <- train %>%
  # creates an unordered factor
  # original variable, dotw, is an ordered factor produced by lubridate (weekdays are ranked)
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
# convert the dotw_simple factor into numbers for the regression model
# model picks one day as reference actegory, with 6 binary (0/1) columns for other days
contrasts(train$dotw_simple) <- contr.treatment(7)

# Now run the model
model1 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation,
  data = train
)

# Summary too long with all station dummies, just show key metrics
cat("Model 1 R-squared:", summary(model1)$r.squared, "\n")
cat("Model 1 Adj R-squared:", summary(model1)$adj.r.squared, "\n")
```

The model uses Monday as the baseline. Each coefficient represents the difference in expected trips per station-hour compared to Monday (i.e. dow_simple = 2 -\> Tuesday)

**Weekday Pattern (Tue-Fri):**

-   All weekdays have positive coefficients (0.029 to 0.052)
-   Tuesday has the highest weekday effect (+0.052)
-   Weekdays likely benefit from concentrated commuting patterns

**Weekend Pattern (Sat-Sun):**

-   Both weekend days have negative coefficients (-0.061 and -0.065)
-   This means FEWER trips per station-hour than Monday

**Hourly Interpretation**

Hour Coefficient Interpretation 0 (baseline) 0.000 trips/hour (midnight) 1 -0.018 slightly fewer than midnight ... 6 +0.151 morning activity starting 7 +0.276 morning rush building 8 +0.487 PEAK morning rush 9 +0.350 post-rush ... 17 +0.568 PEAK evening rush (5 PM!) 18 +0.389 evening declining ... 23 +0.034 late night minimal.

## Model 2: Add Temporal Lags

```{r model2}
model2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day,
  data = train
)


# Summary too long with all station dummies, just show key metrics
cat("Model 2 R-squared:", summary(model2)$r.squared, "\n")
cat("Model 2 Adj R-squared:", summary(model2)$adj.r.squared, "\n")
```

Adding the time lag variables improved R² substantially. This follows from the idea that what happened in the past is often a good proxy for what might happen in the future.

## Model 3: Add Demographics

```{r model3}
model3 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White,
  data = train
)

# Summary too long with all station dummies, just show key metrics
cat("Model 3 R-squared:", summary(model3)$r.squared, "\n")
cat("Model 3 Adj R-squared:", summary(model3)$adj.r.squared, "\n")
```

## Model 4: Add Station Fixed Effects

```{r model4}
model4 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station),
  data = train
)

# Summary too long with all station dummies, just show key metrics
cat("Model 4 R-squared:", summary(model4)$r.squared, "\n")
cat("Model 4 Adj R-squared:", summary(model4)$adj.r.squared, "\n")
```

Station fixed effects capture baseline differences in demand across stations (some are just busier than others!).

## Model 5: Add Rush Hour Interaction

```{r model5}
model5 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day + rush_hour + as.factor(month) +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station) +
    rush_hour * weekend,  # Rush hour effects different on weekends
  data = train
)

cat("Model 5 R-squared:", summary(model5)$r.squared, "\n")
cat("Model 5 Adj R-squared:", summary(model5)$adj.r.squared, "\n")
```

# MODEL EVALUATION

## Calculate Predictions and MAE

```{r calculate_mae}
# Get predictions on test set

# Create day of week factor with treatment (dummy) coding
test <- test %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(test$dotw_simple) <- contr.treatment(7)

test <- test %>%
  mutate(
    pred1 = predict(model1, newdata = test),
    pred2 = predict(model2, newdata = test),
    pred3 = predict(model3, newdata = test),
    pred4 = predict(model4, newdata = test),
    pred5 = predict(model5, newdata = test)
  )

# Calculate MAE for each model
mae_results <- data.frame(
  Model = c(
    "1. Time + Weather",
    "2. + Temporal Lags",
    "3. + Demographics",
    "4. + Station FE",
    "5. + Rush Hour Interaction"
  ),
  MAE = c(
    mean(abs(test$Trip_Count - test$pred1), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred2), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred3), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred4), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred5), na.rm = TRUE)
  )
)

kable(mae_results, 
      digits = 2,
      caption = "Mean Absolute Error by Model (Test Set)",
      col.names = c("Model", "MAE (trips)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Visualize Model Comparison

```{r compare_models}
ggplot(mae_results, aes(x = reorder(Model, -MAE), y = MAE)) +
  geom_col(fill = "#3182bd", alpha = 0.8) +
  geom_text(aes(label = round(MAE, 2)), vjust = -0.5) +
  labs(
    title = "Model Performance Comparison",
    subtitle = "Lower MAE = Better Predictions",
    x = "Model",
    y = "Mean Absolute Error (trips)"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The temporal lag features led to the biggest improvement in the model fit, prompting the greatest decrease in MAE from the baseline model.

# SPACE-TIME ERROR ANALYSIS

## Observed vs. Predicted

We use our best model (Model 2) for error analysis.

```{r obs_vs_pred}
#| message: false
#| warning: false

test <- test %>%
  mutate(
    error = Trip_Count - pred2, # predicted trip count error given by Model 2
    abs_error = abs(error),
    time_of_day = case_when(
      hour < 7 ~ "Overnight",
      hour >= 7 & hour < 10 ~ "AM Rush",
      hour >= 10 & hour < 15 ~ "Mid-Day",
      hour >= 15 & hour <= 18 ~ "PM Rush",
      hour > 18 ~ "Evening"
    )
  )

# Scatter plot by time and day type
ggplot(test, aes(x = Trip_Count, y = pred2)) +
  geom_point(alpha = 0.2, color = "#3182bd") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  facet_grid(weekend ~ time_of_day) +
  labs(
    title = "Observed vs. Predicted Bike Trips",
    subtitle = "Model 2 performance by time period",
    x = "Observed Trips",
    y = "Predicted Trips",
    caption = "Red line = perfect predictions; Green line = actual model fit"
  ) +
  plotTheme
```

The above graphs illustrate that the model often underestimates the trip counts per hour per station. The model struggles especially to accurately predict demand for indego bikes during the peak rush hours of the day.

## Spatial Error Patterns

Now we examine whether prediction errors are clustered in certain parts of Philadelphia.

```{r spatial_errors}
# Calculate MAE by station
station_errors <- test %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

## Create Two Maps Side-by-Side with Proper Legends (sorry these maps are ugly)

# Calculate station errors
station_errors <- test %>%
  filter(!is.na(pred2)) %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs(Trip_Count - pred2), na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE\n(trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),  # Fewer, cleaner breaks
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors",
       subtitle = "Higher in Center City") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand
p2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg\nDemand",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),  # Clear breaks
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand",
       subtitle = "Trips per station-hour") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))




# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE (trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand  
p2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg Demand (trips/hour)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Combine
(p1 | p2)

```

The map on the left-hand side displays spatial clustering of errors, primarily in the neighborhoods in and near center city.

## Temporal Error Patterns

When is the model most wrong?

```{r temporal_errors}
# MAE by time of day and day type
temporal_errors <- test %>%
  group_by(time_of_day, weekend) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(temporal_errors, aes(x = time_of_day, y = MAE, fill = day_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Prediction Errors by Time Period",
    subtitle = "When is the model struggling most?",
    x = "Time of Day",
    y = "Mean Absolute Error (trips)",
    fill = "Day Type"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Errors and Demographics

Are prediction errors related to neighborhood characteristics?

```{r errors_demographics}
#| message: false
#| warning: false

# Join demographic data to station errors
station_errors_demo <- station_errors %>%
  left_join(
    station_attributes %>% select(start_station, Med_Inc, Percent_Taking_Transit, Percent_White),
    by = "start_station"
  ) %>%
  filter(!is.na(Med_Inc))

# Create plots
p1 <- ggplot(station_errors_demo, aes(x = Med_Inc, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Errors vs. Median Income", x = "Median Income", y = "MAE") +
  plotTheme

p2 <- ggplot(station_errors_demo, aes(x = Percent_Taking_Transit, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Transit Usage", x = "% Taking Transit", y = "MAE") +
  plotTheme

p3 <- ggplot(station_errors_demo, aes(x = Percent_White, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Race", x = "% White", y = "MAE") +
  plotTheme

(p1 | p2) / (p3 | plot_spacer())
```

Prediction errors are systematically higher for stations that are in areas with higher shares of the white population, where median income is higher, and where the share of population taking transit is low.

# REPLICATE WITH DIFFERENT QUARTER

## Part 1: **Download data and Adapt Code** for Q4 2024

data from: https://www.rideindego.com/about/data/

```{r load_indego2}
#| message: false
#| warning: false
#| results: hide

# Read Q4 2024 data
indegoq4 <- read_csv(here("data/indego-trips-2024-q4.csv"))

glimpse(indegoq4)

```

```{r parse-datetime}
indegoq4 <- indegoq4 %>%
  # parse datetime
  mutate(
    start_datetime = mdy_hm(start_time),
    end_datetime = mdy_hm(end_time),
    interval60 = floor_date(start_datetime, unit = "hour"), # hourly bins
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7,8,9,16,17,18), 1, 0)
  )
  
 head(indegoq4 %>% select(start_datetime, interval60, week, dotw, hour, weekend))
```

### Check for any data structure changes

```{r date-structure}
# Date range
cat("Date range:", 
    min(mdy_hm(indegoq4$start_time)), "to", 
    max(mdy_hm(indegoq4$start_time)), "\n")

# how many unique stations?
cat("Unique start stations:", length(unique(indegoq4$start_station)), "\n")

# Trip types
table(indegoq4$trip_route_category)

```

```{r load-census}
#| message: false
#| warning: false
#| results: hide

# Get Philadelphia census tracts
philly_censusq4 <- get_acs(
  geography = "tract",
  variables = c(
    "B01003_001",  # Total population
    "B19013_001",  # Median household income
    "B08301_001",  # Total commuters
    "B08301_010",  # Commute by transit
    "B02001_002",  # White alone
    "B25077_001"   # Median home value
  ),
  state = "PA",
  county = "Philadelphia",
  year = 2024,
  geometry = TRUE,
  output = "wide"
) %>%
  rename(
    Total_Pop = B01003_001E,
    Med_Inc = B19013_001E,
    Total_Commuters = B08301_001E,
    Transit_Commuters = B08301_010E,
    White_Pop = B02001_002E,
    Med_Home_Value = B25077_001E
  ) %>%
  mutate(
    Percent_Taking_Transit = (Transit_Commuters / Total_Commuters) * 100,
    Percent_White = (White_Pop / Total_Pop) * 100
  ) %>%
  st_transform(crs = 4326)  #transforming to WGS84 for lat/lon matching, currently in NAD83
```

```{r map-median-income}
#| message: false
#| warning: false

# Map median income
ggplot() +
  geom_sf(data = philly_censusq4, aes(fill = Med_Inc), color = NA) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Context for understanding bike share demand patterns"
  ) +
  # Stations 
  geom_point(
    data = indego,
    aes(x = start_lon, y = start_lat),
    color = "red", size = 0.25, alpha = 0.6
  ) +
  mapTheme

```

```{r spatialjoin-census}

# Create sf object for stations
stations_sfq4 <- indegoq4 %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326)

# Spatial join to get census tract for each station
stations_censusq4 <- st_join(stations_sfq4, philly_censusq4, left = TRUE) %>%
  st_drop_geometry()

# Add back to trip data
indego_censusq4 <- indegoq4 %>%
  left_join(
    stations_censusq4 %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )

# Prepare data for visualization
stations_for_mapq4 <- indegoq4 %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_censusq4 %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Create the map showing problem stations
ggplot() +
  geom_sf(data = philly_censusq4, aes(fill = Med_Inc), color = "white", size = 0.1) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar,
    na.value = "grey90"
  ) +
  # Stations with census data (small grey dots)
  geom_point(
    data = stations_for_mapq4 %>% filter(has_census),
    aes(x = start_lon, y = start_lat),
    color = "grey30", size = 1, alpha = 0.6
  ) +
  # Stations WITHOUT census data (red X marks the spot)
  geom_point(
    data = stations_for_mapq4 %>% filter(!has_census),
    aes(x = start_lon, y = start_lat),
    color = "red", size = 1, shape = 4, stroke = 1.5
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Indego stations shown (RED = no census data match)",
    caption = "Red X marks indicate stations that didn't join to census tracts"
  ) +
  mapTheme
```

```{r missing-data}
#| message: false
#| warning: false

# Identify which stations to keep
valid_stationsq4 <- stations_censusq4 %>%
  filter(!is.na(Med_Inc)) %>%
  pull(start_station)

# Filter trip data to valid stations only
indego_censusq4 <- indegoq4 %>%
  filter(start_station %in% valid_stationsq4) %>%
  left_join(
    stations_censusq4 %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )

```

### Update date ranges for weather data

```{r weather-data}
#| message: false
#| warning: false

# Get weather from Philadelphia International Airport (KPHL)
# This covers Q4 2024: October 1 - December 31
weather_dataq4 <- riem_measures(
  station = "PHL",  # Philadelphia International Airport
  date_start = "2024-10-01",
  date_end = "2024-12-31"
)

# Process weather data
weather_processedq4 <- weather_dataq4 %>%
  mutate(
    interval60 = floor_date(valid, unit = "hour"),
    Temperature = tmpf,  # Temperature in Fahrenheit
    Precipitation = ifelse(is.na(p01i), 0, p01i),  # Hourly precip in inches
    Wind_Speed = sknt  # Wind speed in knots
  ) %>%
  select(interval60, Temperature, Precipitation, Wind_Speed) %>%
  distinct(interval60, .keep_all = TRUE)

# Check for missing hours and interpolate if needed
weather_completeq4 <- weather_processedq4 %>%
  complete(interval60 = seq(min(interval60), max(interval60), by = "hour")) %>%
  fill(Temperature, Precipitation, Wind_Speed, .direction = "down")

# Look at the weather
summary(weather_completeq4 %>% select(Temperature, Precipitation, Wind_Speed))
```

### Create the same 5 models

```{r aggregate-trips-station-hr}
# Count trips by station-hour
trips_panelq4 <- indego_censusq4 %>%
  group_by(interval60, start_station, start_lat, start_lon,
           Med_Inc, Percent_Taking_Transit, Percent_White, Total_Pop) %>%
  summarize(Trip_Count = n()) %>%
  ungroup()

# How many station-hour observations?
nrow(trips_panelq4)

# How many unique stations?
length(unique(trips_panelq4$start_station))

# How many unique hours?
length(unique(trips_panelq4$interval60))

```

```{r expected-panel-sz}

 # Calculate expected panel size
n_stationsq4 <- length(unique(trips_panelq4$start_station))
n_hoursq4 <- length(unique(trips_panelq4$interval60))
expected_rowsq4 <- n_stationsq4 * n_hoursq4

cat("Expected panel rows:", format(expected_rowsq4, big.mark = ","), "\n")

cat("Current rows:", format(nrow(trips_panelq4), big.mark = ","), "\n")

cat("Missing rows:", format(expected_rowsq4 - nrow(trips_panelq4), big.mark = ","), "\n")
```

```{r complete-panel2}
#| message: false
#| warning: false
#| results: hide

# Create complete panel
study_panelq4 <- expand_grid(
  interval60 = unique(trips_panelq4$interval60),
  start_station = unique(trips_panelq4$start_station)
) %>%
  # Join trip counts
  left_join(trips_panelq4 %>% 
              select(interval60, start_station, Trip_Count),
            by = c("interval60", "start_station")) %>%
  # Replace NA trip counts with 0
  mutate(Trip_Count = replace_na(Trip_Count, 0))

glimpse(study_panelq4)
```

```{r add-attributes}
#| message: false
#| warning: false
#| results: hide

# Fill in station attributes (they're the same for all hours)
station_attributesq4 <- trips_panelq4 %>%
  group_by(start_station) %>%
  summarize(
    start_lat = first(start_lat),
    start_lon = first(start_lon),
    Med_Inc = first(Med_Inc),
    Percent_Taking_Transit = first(Percent_Taking_Transit),
    Percent_White = first(Percent_White),
    Total_Pop = first(Total_Pop)
  )

glimpse(station_attributesq4)
```

```{r final-panel}
#| message: false
#| warning: false
#| results: hide

study_panelq4 <- study_panelq4 %>%
  left_join(station_attributesq4, by = "start_station")

glimpse(study_panelq4)

# Verify we have complete panel
cat("Complete panel rows:", format(nrow(study_panelq4), big.mark = ","), "\n")
```

```{r add-time-features}
#| message: false
#| warning: false
#| results: hide


study_panelq4 <- study_panelq4 %>%
  mutate(
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

glimpse(study_panelq4)
```

```{r add-weather}
#| message: false
#| warning: false
#| results: hide

study_panelq4 <- study_panelq4 %>%
  left_join(weather_completeq4, by = "interval60")

glimpse(study_panelq4)

summary(study_panelq4 %>% select(Trip_Count, Temperature, Precipitation))


```

```{r add-temporal-lags}
#| message: false
#| warning: false

# Sort by station and time
study_panelq4 <- study_panelq4 %>%
  arrange(start_station, interval60)

# Create lag variables WITHIN each station
study_panelq4 <- study_panelq4 %>%
  group_by(start_station) %>%
  mutate(
    lag1Hour = lag(Trip_Count, 1),
    lag2Hours = lag(Trip_Count, 2),
    lag3Hours = lag(Trip_Count, 3),
    lag12Hours = lag(Trip_Count, 12),
    lag1day = lag(Trip_Count, 24)
  ) %>%
  ungroup()

# Remove rows with NA lags (first 24 hours for each station)
study_panel_completeq4 <- study_panelq4 %>%
  filter(!is.na(lag1day))

cat("Rows after removing NA lags:", format(nrow(study_panel_completeq4), big.mark = ","), "\n")
```

```{r train-test-split}
#| message: false
#| warning: false

# Split by week
# Q1 has weeks 1-13 (Oct-Dec)
# Train on weeks 1-9 (Oct 1 - early December)
# Test on weeks 10-13 (rest of December)

# Which stations have trips in BOTH early and late periods?
early_stationsq4 <- study_panel_completeq4 %>%
  filter(week < 49) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

late_stationsq4 <- study_panel_completeq4 %>%
  filter(week >= 49) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

# Keep only stations that appear in BOTH periods
common_stationsq4 <- intersect(early_stationsq4, late_stationsq4)


# Filter panel to only common stations
study_panel_completeq4 <- study_panel_completeq4 %>%
  filter(start_station %in% common_stationsq4)

# NOW create train/test split
trainq4 <- study_panel_completeq4 %>%
  filter(week < 49)

testq4 <- study_panel_completeq4 %>%
  filter(week >= 49)

cat("Training observations:", format(nrow(trainq4), big.mark = ","), "\n")

cat("Testing observations:", format(nrow(testq4), big.mark = ","), "\n")

```

```{r Model-1}
#| message: false
#| warning: false

# Create day of week factor with treatment (dummy) coding
trainq4 <- trainq4 %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(trainq4$dotw_simple) <- contr.treatment(7)

# Now run the model
q4model1 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation,
  data = trainq4
)

cat("Model 1 R-squared:", summary(q4model1)$r.squared, "\n")
cat("Model 1 Adj R-squared:", summary(q4model1)$adj.r.squared, "\n")

```

```{r Model-2}

# With temporal lags
q4model2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation + lag1Hour + lag3Hours + lag1day,
  data = trainq4
)

cat("Model 2 R-squared:", summary(q4model2)$r.squared, "\n")
cat("Model 2 Adj R-squared:", summary(q4model2)$adj.r.squared, "\n")

```

```{r Model-3}

# With demographics
q4model3 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation + lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White,
  data = trainq4
)

cat("Model 3 R-squared:", summary(q4model3)$r.squared, "\n")
cat("Model 3 Adj R-squared:", summary(q4model3)$adj.r.squared, "\n")
```

```{r Model-4}
q4model4 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station),
  data = trainq4
)

# Summary too long with all station dummies, just show key metrics
cat("Model 4 R-squared:", summary(q4model4)$r.squared, "\n")
cat("Model 4 Adj R-squared:", summary(q4model4)$adj.r.squared, "\n")
```

```{r Model-5}

q4model5 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day + rush_hour + as.factor(month) +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station) +
    rush_hour * weekend,  # Rush hour effects different on weekends
  data = trainq4
)

cat("Model 5 R-squared:", summary(q4model5)$r.squared, "\n")
cat("Model 5 Adj R-squared:", summary(q4model5)$adj.r.squared, "\n")
```

### Calculate MAE for each model

```{r calculate-MAE}

# Get predictions on test set

# Create day of week factor with treatment (dummy) coding
testq4 <- testq4 %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(testq4$dotw_simple) <- contr.treatment(7)

testq4 <- testq4 %>%
  mutate(
    pred1 = predict(q4model1, newdata = testq4),
    pred2 = predict(q4model2, newdata = testq4),
    pred3 = predict(q4model3, newdata = testq4),
    pred4 = predict(q4model4, newdata = testq4),
    pred5 = predict(q4model5, newdata = testq4)
  )

# Calculate MAE for each model
mae_resultsq4 <- data.frame(
  Model = c(
    "1. Time + Weather",
    "2. + Temporal Lags",
    "3. + Demographics",
    "4. + Station FE",
    "5. + Rush Hour Interaction"
  ),
  MAE = c(
    mean(abs(testq4$Trip_Count - testq4$pred1), na.rm = TRUE),
    mean(abs(testq4$Trip_Count - testq4$pred2), na.rm = TRUE),
    mean(abs(testq4$Trip_Count - testq4$pred3), na.rm = TRUE),
    mean(abs(testq4$Trip_Count - testq4$pred4), na.rm = TRUE),
    mean(abs(testq4$Trip_Count - testq4$pred5), na.rm = TRUE)
  )
)

kable(mae_resultsq4, 
      digits = 2,
      caption = "Mean Absolute Error by Model (Test Set)",
      col.names = c("Model", "MAE (trips)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### **Compare results** to Q1 2025

```{r compare-MAE-Q1-Q4}

# Create side-by-side bar chart with Q1 and Q4 MAE's for each model
# Q1
maeq1 <- ggplot(mae_results, aes(x = reorder(Model, -MAE), y = MAE)) +
  geom_col(fill = "#3182bd", alpha = 0.8) +
  geom_text(aes(label = round(MAE, 2)), vjust = -0.5) +
  coord_cartesian(ylim = c(0, 0.75)) +
  labs(
    title = "Model Performance Comparison",
    subtitle = "Lower MAE = Better Predictions",
    x = "Model",
    y = "Mean Absolute Error (trips)"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# Q4
maeq4 <- ggplot(mae_resultsq4, aes(x = reorder(Model, -MAE), y = MAE)) +
  geom_col(fill = "51355a", alpha = 0.8) +
  geom_text(aes(label = round(MAE, 2)), vjust = -0.5) +
  coord_cartesian(ylim = c(0, 0.75)) +
  labs(
    title = "Model Performance Comparison",
    subtitle = "Lower MAE = Better Predictions",
    x = "Model",
    y = "Mean Absolute Error (trips)"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

(maeq1 | maeq4)
```

The 2024 Q4 models have lower MAEs for each model type than the models built from 2025 Q1. The models based on Q1 and Q4, both see the biggest improvement in MAE once the temporal lag variables are added into their respective Model 2. As such, the best performing model for predictions based on Q1 and Q4 data is Model 2, +Temporal Lags.

The MAE values for each model built from either Q1 or Q4 data might differ if all bike stations during one period of the year are used at a higher rate compared to another. A higher and more consistent rate of bike usage across all stations provides the model with more observations to train itself on, and thus produce slightly more accurate predictions. For example, Q1 may have observations of bike trips only during winter months when bike usage is overall decreased due to the cold and less consistent across all bike stations, while Q4 holds observations of bike trips that occurred during the transition to winter, when weather might be warmer, and thus trips are more frequently taken consistently across stations.

```{r compare-temporal-patterns}

# Side-by-side barchart graph for Q1 and Q4 best models (Model 2, +Temporal Lags) MAEs for weekends and weekdays
# MAE by time of day and day type - Q1, Model 2
temporal_errors <- test %>%
  group_by(time_of_day, weekend) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

temperrq1 <- ggplot(temporal_errors, aes(x = time_of_day, y = MAE, fill = day_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Prediction Errors by Time Period \n Q1 2025",
    subtitle = "When is the model struggling most?",
    x = "Time of Day",
    y = "Mean Absolute Error (trips)",
    fill = "Day Type"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# MAE by time of day and day type - Q4, Model 2
testq4 <- testq4 %>%
  mutate(
    error = Trip_Count - pred2, # predicted trip count error given by Model 2
    abs_error = abs(error),
    time_of_day = case_when(
      hour < 7 ~ "Overnight",
      hour >= 7 & hour < 10 ~ "AM Rush",
      hour >= 10 & hour < 15 ~ "Mid-Day",
      hour >= 15 & hour <= 18 ~ "PM Rush",
      hour > 18 ~ "Evening"
    )
  )

temporal_errorsq4 <- testq4 %>%
  group_by(time_of_day, weekend) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

temperrq4 <- ggplot(temporal_errorsq4, aes(x = time_of_day, y = MAE, fill = day_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Prediction Errors by Time Period \n Q4 2024",
    subtitle = "When is the model struggling most?",
    x = "Time of Day",
    y = "Mean Absolute Error (trips)",
    fill = "Day Type"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

(temperrq1 | temperrq4)

```

The two bar charts of the Q1 and Q4 Model 2 Temporal MAE values by day type substantiates that Q1's Model 2 is generally weaker than Q4's Model 2 which has lower MAE values for each time period during the day and across weekend and weekday day types. They are similar to one another, however, in that each model better predicts trips on weekends rather than weekdays for almost every time period of the day, except for 'Mid-Day', for which Q4's Model 2 predicts trip counts with the same average accuracy for both weekend and weekday.

## Part 2: Error Analysis

Analyze the model's errors in detail:

1.  **Spatial patterns:**

    ```{r MAE-by-station}
    #| message: false
    #| warning: false

    ## Create Two Maps Side-by-Side with Proper Legends

    # Calculate station errors
    station_errorsq4 <- testq4 %>%
      filter(!is.na(pred2)) %>%
      group_by(start_station, start_lat, start_lon) %>%
      summarize(
        MAE = mean(abs(Trip_Count - pred2), na.rm = TRUE),
        avg_demand = mean(Trip_Count, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      filter(!is.na(start_lat), !is.na(start_lon))

    # Map 1: Prediction Errors
    p1 <- ggplot() +
      geom_sf(data = philly_censusq4, fill = "grey95", color = "white", size = 0.2) +
      geom_point(
        data = station_errorsq4,
        aes(x = start_lon, y = start_lat, color = MAE),
        size = 3.5,
        alpha = 0.7
      ) +
      scale_color_viridis(
        option = "plasma",
        name = "MAE\n(trips)",
        direction = -1,
        breaks = c(0.5, 1.0, 1.5),  # Fewer, cleaner breaks
        labels = c("0.5", "1.0", "1.5")
      ) +
      labs(title = "Prediction Errors",
           subtitle = "Higher in Center City") +
      mapTheme +
      theme(
        legend.position = "right",
        legend.title = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 9),
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 10)
      ) +
      guides(color = guide_colorbar(
        barwidth = 1.5,
        barheight = 12,
        title.position = "top",
        title.hjust = 0.5
      ))

    # Map 2: Average Demand
    p2 <- ggplot() +
      geom_sf(data = philly_censusq4, fill = "grey95", color = "white", size = 0.2) +
      geom_point(
        data = station_errorsq4,
        aes(x = start_lon, y = start_lat, color = avg_demand),
        size = 3.5,
        alpha = 0.7
      ) +
      scale_color_viridis(
        option = "viridis",
        name = "Avg\nDemand",
        direction = -1,
        breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),  # Clear breaks
        labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
      ) +
      labs(title = "Average Demand",
           subtitle = "Trips per station-hour") +
      mapTheme +
      theme(
        legend.position = "right",
        legend.title = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 9),
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 10)
      ) +
      guides(color = guide_colorbar(
        barwidth = 1.5,
        barheight = 12,
        title.position = "top",
        title.hjust = 0.5
      ))

    # Combine
    (p1 | p2)
    ```

Just like the Model 2 based on Q1 data, the Model 2 based on Q4 data produces higher prediction errors in neighborhoods in and near center city. One reason why this could be the case is that Model 2 does not contain some key variables, unique to downtown areas, which influence bike trip incidence more than in other parts of the city.

1.  **Temporal patterns:**

    ```{r observed-v-predicted}
    #| message: false
    #| warning: false

    testq4 <- testq4 %>%
      mutate(
        error = Trip_Count - pred2,
        abs_error = abs(error),
        time_of_day = case_when(
          hour < 7 ~ "Overnight",
          hour >= 7 & hour < 10 ~ "AM Rush",
          hour >= 10 & hour < 15 ~ "Mid-Day",
          hour >= 15 & hour <= 18 ~ "PM Rush",
          hour > 18 ~ "Evening"
        )
      )

    # Scatter plot by time and day type
    ggplot(testq4, aes(x = Trip_Count, y = pred2)) +
      geom_point(alpha = 0.2, color = "#3182bd") +
      geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1) +
      geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
      facet_grid(weekend ~ time_of_day) +
      labs(
        title = "Observed vs. Predicted Bike Trips",
        subtitle = "Model 2 performance by time period",
        x = "Observed Trips",
        y = "Predicted Trips",
        caption = "Red line = perfect predictions; Green line = actual model fit"
      ) +
      plotTheme
    ```

    ```{r temporal-patterns}
    # MAE by time of day and day type
    temporal_errorsq4 <- testq4 %>%
      group_by(time_of_day, weekend) %>%
      summarize(
        MAE = mean(abs_error, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

    ggplot(temporal_errorsq4, aes(x = time_of_day, y = MAE, fill = day_type)) +
      geom_col(position = "dodge") +
      scale_fill_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
      labs(
        title = "Prediction Errors by Time Period",
        subtitle = "When is the model struggling most?",
        x = "Time of Day",
        y = "Mean Absolute Error (trips)",
        fill = "Day Type"
      ) +
      plotTheme +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
    ```

Q4's Model 2 produces the greatest prediction errors mostly on weekdays, except for the 'Mid-day' and 'Overnight' times of day, but only marginally so. This means that Model 2 is less accurate at predicting trip counts during weekdays, but improves during the middle of the day and overnight in comparison to its performance on the weekends.

1.  **Demographic patterns:**

    ```{r demo-patterns}
    #| message: false
    #| warning: false

    # Join demographic attributes to station errors
    station_errors_demoq4 <- station_errorsq4 %>%
      left_join(
        station_attributesq4 %>% select(start_station, Med_Inc, Percent_Taking_Transit, Percent_White),
        by = "start_station"
      ) %>%
      filter(!is.na(Med_Inc))

    # Create plots
    p1 <- ggplot(station_errors_demoq4, aes(x = Med_Inc, y = MAE)) +
      geom_point(alpha = 0.5, color = "#3182bd") +
      geom_smooth(method = "lm", se = FALSE, color = "red") +
      scale_x_continuous(labels = scales::dollar) +
      labs(title = "Errors vs. Median Income", x = "Median Income", y = "MAE") +
      plotTheme

    p2 <- ggplot(station_errors_demoq4, aes(x = Percent_Taking_Transit, y = MAE)) +
      geom_point(alpha = 0.5, color = "#3182bd") +
      geom_smooth(method = "lm", se = FALSE, color = "red") +
      labs(title = "Errors vs. Transit Usage", x = "% Taking Transit", y = "MAE") +
      plotTheme

    p3 <- ggplot(station_errors_demoq4, aes(x = Percent_White, y = MAE)) +
      geom_point(alpha = 0.5, color = "#3182bd") +
      geom_smooth(method = "lm", se = FALSE, color = "red") +
      labs(title = "Errors vs. Race", x = "% White", y = "MAE") +
      plotTheme

    (p1 | p2) / (p3 | plot_spacer())
    ```

-   Are certain communities systematically harder to predict? Yes
-   What are the equity implications? ??

## Part 3: Feature Engineering & model improvement

Add 2-3 NEW features to improve the model: UPenn School Sessions, Daylight hours (sunrise/sunset), Wind Chill levels, and the Distance to center city

The new temporal feature, upenn_in_session, was added to try to capture the patterns of higher or lower activity in the west philly area linked to school sessions. This activity could possibly influence the trip counts in the relevant areas and therefore is a valid feature to include in the model. Similarly, daylight hours is also a reasonable predictor of bike demand as whether it is light out or dark could directly influence people's choice to begin a bike trip. Wind chill levels are additional feature that captures one specific element of weather conditions. If wind chill levels are higher, people may be less likely to want to ride an indego bike due to greater discomfort. Lastly, a variable that addresses each station-trip's distance to center city could be useful as this could be a key factor in one's decision to begin a bike trip. If the station is closer to downtown, one's trip to their destination in the city's hub of activity may be shorter and therefore, more desirable, as a shorter bike trip is cheaper.

```{r temporal-features-new}
#| message: false
#| warning: false

# Create 1st temporal feature that categorizes the time when UPenn is in session
study_panel_completeq4 <- study_panel_completeq4 %>% 
  mutate(
    upenn_in_session = case_when(
      # Fall semester in session
      date >= as.Date("2024-10-01") & date <= as.Date("2024-11-27") ~ 1,
      # Finals/end of semester
      date >= as.Date("2024-12-02") & date <= as.Date("2024-12-20") ~ 1,
      # Everything else = not in session (Thanksgiving + Winter break) 
      TRUE ~ 0
    )
  )

# Create 2nd temporal feature: daylight hours
# Using 'suncalc' package

# Get unique dates in panel
# Create dataframe with row for each unique date in Q4 (92 days), n=92, each with same lat/lon for Philadelphia area
dates_q4 <- data.frame(
  # parse unique dates from panel (has many rows with same date)
  date = unique(study_panel_completeq4$date), 
  lat = 39.9526,   # Philadelphia latitude
  lon = -75.1652   # Philadelphia longitude
)
  
# Calculate sunset/sunrise
sun_times <- getSunlightTimes(
  data = dates_q4,
  # function only return sunrise and sunset
  keep = c("sunrise", "sunset"), 
  tz = "America/New York"
) %>% 
  mutate(
    # difftime - subtracts sunrise from sunset to get length of day
    # units = "hours" - returns result in hours
    # as.numeric strips "hours" label so it is plain number - to   prevent modeling issues
    daylight_hours = as.numeric(difftime(sunset, sunrise, units = "hours")) 
  ) %>% 
  # drop lat and lon columns - no longer need
  # keep date to join back to complete panel
  select(date, daylight_hours)

# Join daylight feature to complete panel
study_panel_completeq4 <- study_panel_completeq4 %>% 
  left_join(sun_times, by = "date")
```

```{r weather-feature-new}
#| message: false
#| warning: false

# Create wind chill feature
weather_completeq4 <- weather_completeq4 %>% 
  mutate(
    #Only calculate wind chill when below 50 F with wind > 3mph
    wind_chill = case_when(
      Temperature <= 50 & Wind_Speed > 3 ~ 
        35.74 + (0.6215 * Temperature) - (35.75*Wind_Speed^0.16) +
        (0.4275 * Temperature * Wind_Speed^0.16),
      # otherwise just use actual temperature,
      TRUE ~ Temperature
    )
  )

# Join wind chill to complete panel
study_panel_completeq4 <- study_panel_completeq4 %>% 
  left_join(weather_completeq4, by = "interval60")
```

```{r spatial-feature}
#| message: false
#| warning: false

# Create distance to center city feature
# load center city boundary 
center_city <- st_read("data/center_city_boundary.geojson")

# transforming to appropriate CRS
center_city <- st_transform(center_city, 3365)

# join to complete panel
study_panel_completeq4 <- study_panel_completeq4 %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326, remove = FALSE) %>%
  st_transform(crs = 3365) %>%
  mutate(
    dist_centercity = as.numeric(st_distance(., st_union(center_city)))
  ) %>% 
  st_drop_geometry()
```

**Implementation:**

```{r Model-with-new-features}
#| message: false
#| warning: false

# re-create train/test split with updated complete panel
trainq4 <- study_panel_completeq4 %>%
  filter(week < 49)

testq4 <- study_panel_completeq4 %>%
  filter(week >= 49)

# Create day of week factor with treatment (dummy) coding
trainq4 <- trainq4 %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(trainq4$dotw_simple) <- contr.treatment(7)

# cleaning up duplicate columns
trainq4 <- trainq4 %>%
  rename(
    Temperature = Temperature.x,
    Precipitation = Precipitation.x,
    Wind_Speed = Wind_Speed.x
  ) %>%
  select(-ends_with(".y"))  # drop the duplicate .y column

# Model 2 with new features
q4model6 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation + lag1Hour + lag3Hours + lag1day + upenn_in_session + daylight_hours + wind_chill + dist_centercity,
  data = trainq4
)

summary(q4model6)
```

```{r calculate-new-MAE}
#| message: false
#| warning: false

# Get predictions on test set

# Create day of week factor with treatment (dummy) coding
testq4 <- testq4 %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(testq4$dotw_simple) <- contr.treatment(7)

# cleaning up duplicate columns
testq4 <- testq4 %>%
  rename(
    Temperature = Temperature.x,
    Precipitation = Precipitation.x,
    Wind_Speed = Wind_Speed.x
  ) %>%
  select(-ends_with(".y"))  # drop the duplicate .y column

testq4 <- testq4 %>%
  mutate(
    pred1 = predict(q4model1, newdata = testq4),
    pred2 = predict(q4model2, newdata = testq4),
    pred3 = predict(q4model3, newdata = testq4),
    pred4 = predict(q4model4, newdata = testq4),
    pred5 = predict(q4model5, newdata = testq4),
    pred6 = predict(q4model6, newdata = testq4)
  )

# Calculate MAE for each model
mae_resultsq4 <- data.frame(
  Model = c(
    "1. Time + Weather",
    "2. + Temporal Lags",
    "3. + Demographics",
    "4. + Station FE",
    "5. + Rush Hour Interaction",
    "6. + New Features"
  ),
  MAE = c(
    mean(abs(testq4$Trip_Count - testq4$pred1), na.rm = TRUE),
    mean(abs(testq4$Trip_Count - testq4$pred2), na.rm = TRUE),
    mean(abs(testq4$Trip_Count - testq4$pred3), na.rm = TRUE),
    mean(abs(testq4$Trip_Count - testq4$pred4), na.rm = TRUE),
    mean(abs(testq4$Trip_Count - testq4$pred5), na.rm = TRUE),
    mean(abs(testq4$Trip_Count - testq4$pred6), na.rm = TRUE)
  )
)

kable(mae_resultsq4, 
      digits = 2,
      caption = "Mean Absolute Error by Model (Test Set)",
      col.names = c("Model", "MAE (trips)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

Model 6 does not produce a lower MAE (0.45) than the simpler Model 2 that has an MAE of 0.44 bike trips. Thus, the newly added features did not improve the best performing model, Model 2, with temporal lags.

### Did they improve predictions? Where?

```{r}
# Calculate station errors
station_errorsq4 <- testq4 %>%
  filter(!is.na(pred6)) %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs(Trip_Count - pred6), na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_censusq4, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errorsq4,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE\n(trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),  # Fewer, cleaner breaks
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors",
       subtitle = "Higher in Center City") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand
p2 <- ggplot() +
  geom_sf(data = philly_censusq4, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errorsq4,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg\nDemand",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),  # Clear breaks
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand",
       subtitle = "Trips per station-hour") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))

# Combine
(p1 | p2)
```

```{r}

# Look at top 10 stations that the model has difficulty predicting
top10 <- station_errorsq4 %>% 
  select(start_station, MAE, avg_demand, start_lat, start_lon) %>% 
  arrange(desc(MAE)) %>% 
  head(10)

# Where are they?
p1 <- ggplot() +
  geom_sf(data = philly_censusq4, fill = "grey90", color = "white", size = 0.1) +
  geom_point(
    data = top10,
    aes(x = start_lon, y = start_lat),
    color = "red", size = 3
  ) +
  labs(
    title = "Stations with Highest Prediction Error",
    subtitle = "Top 10 stations by Mean Absolute Error",
    caption = "Red points indicate stations where model performed worst"
  ) +
  mapTheme
p1
```

```{r}
#| message: false
#| warning: false

testq4 <- testq4 %>%
  mutate(
    error = Trip_Count - pred6,
    abs_error = abs(error),
    time_of_day = case_when(
      hour < 7 ~ "Overnight",
      hour >= 7 & hour < 10 ~ "AM Rush",
      hour >= 10 & hour < 15 ~ "Mid-Day",
      hour >= 15 & hour <= 18 ~ "PM Rush",
      hour > 18 ~ "Evening"
    )
  )

# Scatter plot by time and day type
ggplot(testq4, aes(x = Trip_Count, y = pred6)) +
  geom_point(alpha = 0.2, color = "#3182bd") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  facet_grid(weekend ~ time_of_day) +
  labs(
    title = "Observed vs. Predicted Bike Trips",
    subtitle = "Model 6 performance by time period",
    x = "Observed Trips",
    y = "Predicted Trips",
    caption = "Red line = perfect predictions; Green line = actual model fit"
  ) +
  plotTheme
```

**Try a poisson model for count data**

```{r poisson}
#| message: false
#| warning: false

poisson <- glm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation + lag1Hour + lag3Hours + lag1day + upenn_in_session + daylight_hours + wind_chill + dist_centercity,
  data = trainq4,
  family = "poisson"
)

# Summary
summary(poisson)
```

```{r poisson-model6-compare}
# Add poisson predictions to test set
testq4 <- testq4 %>%
  mutate(
    pred_poisson = predict(poisson, newdata = testq4, type = "response")
  )

# Compare MAE
mae_linear <- mean(abs(testq4$Trip_Count - testq4$pred6), na.rm = TRUE)
mae_poisson <- mean(abs(testq4$Trip_Count - testq4$pred_poisson), na.rm = TRUE)

cat("Linear Model 6 MAE:", round(mae_linear, 3), "\n")
cat("Poisson Model MAE:", round(mae_poisson, 3), "\n")
```

The Poisson model has a lower MAE than the linear Model 6, meaning that the Poisson regression produces a better model fit than simple linear regression, generating more accurate trip count predictions per station-hour.

## Part 4: Critical Reflection

**Operational Implications**

Model 2 (MAE = 0.44 trips/station-hour) is the best performing model and accurate enough for broad rebalancing decisions given average demand is well below 2 trips/hour. However, the model consistently underpredicts during AM and PM weekday rush, often the highest-stakes time for rebalancing. Adding new features (Model 6) improved on Models 3–5 but did not beat Model 2, suggesting temporal lags are the most important predictor. It is recommended to use this model for overnight rebalancing planning, but suggested to flag high-demand stations for manual oversight during peak hours.

**Equity Considerations**

Prediction errors are spatially concentrated in Center City and surrounding areas, suggesting that the most used stations are also the hardest to predict. Lower errors in lower-income neighborhoods likely reflect very low trip volumes rather than true model accuracy. A model-driven rebalancing system could thus systematically under-supply high-demand stations, but continue appearing unbiased on average. Therefore, for rebalancing, this model would be best supplemented by looking specifically at the prediction errors of high-demand stations as this would differ from the prediction error for the average station.

**Model Limitations**

The final model likely misses irregular demand spikes from events like concerts or sports games as it does not contain any variable that captures these one-off moments. The model, as it stands, assumes that the relationship between the predictors and trip counts will stay consistent over time. However, this assumption may not hold as ridership habits shift or new stations are added to the network. With more time and data, it would be interesting to incorporate an event calendar for major Philadelphia venues, train across multiple quarters to better capture seasonal patterns, and explore whether proximity to specific destinations like transit stops or universities could further improve predictions beyond distance to Center City alone.

------------------------------------------------------------------------
