---
title: "Lab 0: Getting Started with dplyr"
subtitle: "My First Data Analysis"
author: "Maude Ceruso"
date: 01-20-2026
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

# Overview

Welcome to your first lab! In this (not graded) assignment, you'll
practice the fundamental dplyr operations I overviewed in class using
car sales data. This lab will help you get comfortable with:

-   Basic data exploration
-   Column selection and manipulation
-   Creating new variables
-   Filtering data
-   Grouping and summarizing
-   

# Setup

```{r setup}
# Load the tidyverse library
library(tidyverse)

# Read in the car sales data
# Make sure the data file is in your lab_0/data/ folder
car_data <- read_csv("data/car_sales_data.csv")
```

# Exercise 1: Getting to Know Your Data

## 1.1 Data Structure Exploration

Explore the structure of your data and answer these questions:

```{r explore-structure}
#| message: false
#| warning: false

# Use glimpse() to see the data structure
glimpse(car_data)

# Check the column names
names(car_data)

# Look at the first few rows
head(car_data)
```

**Questions to answer:** - How many rows and columns does the dataset
have? - What types of variables do you see (numeric, character, etc.)? -
Are there any column names that might cause problems? Why?

**Your answers:**

\- Rows: 50000 - Columns: 7\
- Variable types: numeric and character variables - Problematic names:
"Engine size" or "Fuel type" due to spaces in between words

## 1.2 Tibble vs Data Frame

Compare how tibbles and data frames display:

```{r tibble-vs-dataframe}
#| message: false
#| warning: false

# Look at the tibble version (what we have)
print(car_data)

# Convert to regular data frame and display
car_df <- as.data.frame(car_data)
print(car_df)
```

**Question:** What differences do you notice in how they print?

**Your answer:** The tibble shows the first 10 rows of data, and each
variable is denoted as either character or numeric, whereas the
dataframe shows rows until it reaches its max, in this case 142 rows.
The dataframe does not specify whether the variables are characters or
numeric.

# Exercise 2: Basic Column Operations

## 2.1 Selecting Columns

```{r column-selection}
#| message: false
#| warning: false

# Select just Model and Mileage columns
select(car_data, Model, Mileage)

# Select Manufacturer, Price, and Fuel type

select(car_data, Manufacturer, Price, 'Fuel type')

# Challenge: Select all columns EXCEPT Engine Size

select(car_data, -'Engine size')
```

## 2.2 Renaming Columns

```{r rename-columns}
#| message: false
#| warning: false

# Rename 'Year of manufacture' to year
rename(car_data, year = 'Year of manufacture')

car_data <- car_data %>%
  rename(Year = 'Year of manufacture')

# Check that it worked
names(car_data)
```

**Question:** Why did we need backticks around `Year of manufacture` but
not around `year`?

**Your answer:** Due to the spaces between each word.

# Exercise 3: Creating New Columns

## 3.1 Calculate Car Age

```{r calculate-age}
#| message: false
#| warning: false

# Create an 'age' column (2025 minus year of manufacture)
car_data <- car_data %>%
  mutate(Age = 2025 - Year)

# Create a mileage_per_year column  
car_data_new <- car_data %>%
  mutate(mileage_per_year = Mileage/Year)

# Look at your new columns

print(car_data_new)

select(car_data_new, Model, Year, Age, Mileage, mileage_per_year)

```

## 3.2 Categorize Cars

```{r categorize-cars}
#| message: false
#| warning: false

# Create a price_category column where if price is < 15000, its is coded as budget, between 15000 and 30000 is midrange and greater than 30000 is mid-range (use case_when)

car_data_new <- car_data_new %>%
  mutate(price_category = case_when(
    Price < 15000 ~ "Budget",
    Price <= 30000 ~ "Midrange",
    Price > 30000 ~ "Expensive",
    TRUE ~ NA_character_
  ))

# Check your categories select the new column and show it
select(car_data_new, price_category)
```

# Exercise 4: Filtering Practice

## 4.1 Basic Filtering

```{r basic-filtering}
#| message: false
#| warning: false

# Find all Toyota cars
filter(car_data_new, Manufacturer == "Toyota")

# Find cars with mileage less than 30,000
filter(car_data_new, Mileage < 30000)

# Find luxury cars (from price category) with low mileage
filter(car_data_new, price_category == "Expensive" & Mileage < 30000)

```

## 4.2 Multiple Conditions

```{r multiple-conditions}
#| message: false
#| warning: false

# Find cars that are EITHER Honda OR Nissan
filter(car_data_new, Manufacturer %in% c("Honda","Nissan"))

# Find cars with price between $20,000 and $35,000
filter(car_data_new, Price >= 20000 & Price <= 35000)

# Find diesel cars less than 10 years old
filter(car_data_new, `Fuel type` == "Diesel" & Age < 10 )

```

**Question:** How many diesel cars are less than 10 years old?

**Your answer:** 2,040

# Exercise 5: Grouping and Summarizing

## 5.1 Basic Summaries

```{r basic-summaries}
#| message: false
#| warning: false

# Calculate average price by manufacturer
avg_price_by_brand <- car_data_new %>%
  group_by(Manufacturer) %>%
  summarize(avg_price = mean(Price, na.rm = TRUE))

avg_price_by_brand

# Calculate average mileage by fuel type

avg_mil_by_fuel <- car_data_new %>%
  group_by(`Fuel type`) %>%
  summarize(avg_mil = mean(Mileage, na.rm = TRUE))

avg_mil_by_fuel

# Count cars by manufacturer

count_car_by_mfctr <- car_data_new %>%
  group_by(Manufacturer) %>%
  summarize(count = n ())
count_car_by_mfctr
```

## 5.2 Categorical Summaries

```{r categorical-summaries}
#| message: false
#| warning: false

# Frequency table for price categories

price_freq_tbl <- car_data_new %>%
  group_by(price_category) %>%
  summarize(count = n())

price_freq_tbl
```
