---
title: "Lab 1: Census Data Quality for Policy Decisions"
subtitle: "Evaluating Data Reliability for Algorithmic Decision-Making"
author: "Maude Ceruso"
date: February 1, 2026
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

# Assignment Overview

## Scenario

You are a data analyst for the **\[Your State\] Department of Human
Services**. The department is considering implementing an algorithmic
system to identify communities that should receive priority for social
service funding and outreach programs. Your supervisor has asked you to
evaluate the quality and reliability of available census data to inform
this decision.

Drawing on our Week 2 discussion of algorithmic bias, you need to assess
not just what the data shows, but how reliable it is and what
communities might be affected by data quality issues.

## Learning Objectives

-   Apply dplyr functions to real census data for policy analysis
-   Evaluate data quality using margins of error
-   Connect technical analysis to algorithmic decision-making
-   Identify potential equity implications of data reliability issues
-   Create professional documentation for policy stakeholders

## Submission Instructions

**Submit by posting your updated portfolio link on Canvas.** Your
assignment should be accessible at `your-portfolio-url/labs/lab_1/`

Make sure to update your `_quarto.yml` navigation to include this
assignment under an "Labs" menu.

# Part 1: Portfolio Integration

Create this assignment in your portfolio repository under an
`labs/lab_1/` folder structure. Update your navigation menu to include:

```         
- text: Assignments
  menu:
    - href: labs/lab_1/your_file_name.qmd
      text: "Lab 1: Census Data Exploration"
```

If there is a special character like a colon, you need use double quote
mark so that the quarto can identify this as text

# Setup

```{r setup}
# Load required packages (hint: you need tidycensus, tidyverse, and knitr)
library(tidycensus)
library(tidyverse)
library(knitr)

# Set your Census API key
census_api_key("c61294d533815353a666f356be78c2a75d0b1f00")

# Choose your state for analysis - assign it to a variable called my_state
my_state <- "New York"
```

**State Selection:** I have chosen **New York** for this analysis
because: this is my home state.

# Part 2: County-Level Resource Assessment

## 2.1 Data Retrieval

**Your Task:** Use `get_acs()` to retrieve county-level data for your
chosen state.

**Requirements:** - Geography: county level - Variables: median
household income (B19013_001) and total population (B01003_001)\
- Year: 2022 - Survey: acs5 - Output format: wide

**Hint:** Remember to give your variables descriptive names using the
`variables = c(name = "code")` syntax.

```{r county-data}
# Write your get_acs() code here
state_data <- get_acs(
  geography = "county",
  variables = c(medHHinc = "B19013_001", totpop = "B01003_001"),
  year = 2022,
  survey = "acs5",
  output = "wide",
  state = my_state
)

# Clean the county names to remove state name and "County" 
# Hint: use mutate() with str_remove()
ny_clean <- state_data %>% 
  mutate(
    #remove state name form county names
    county_name = str_remove(NAME, ", New York"),
    #remove "County" word
    county_name = str_remove(county_name, " County")
  )

# Display the first few rows
head(ny_clean)
select(ny_clean, NAME, county_name)
```

## 2.2 Data Quality Assessment

**Your Task:** Calculate margin of error percentages and create
reliability categories.

**Requirements:** - Calculate MOE percentage: (margin of error /
estimate) \* 100 - Create reliability categories: - High Confidence: MOE
\< 5% - Moderate Confidence: MOE 5-10%\
- Low Confidence: MOE \> 10% - Create a flag for unreliable estimates
(MOE \> 10%)

**Hint:** Use `mutate()` with `case_when()` for the categories.

```{r income-reliability}
# Calculate MOE percentage and reliability categories using mutate()
ny_reliability <- ny_clean %>%
  mutate(moe_percentage = round((medHHincM / medHHincE)* 100, 2),
         reliability = case_when(
           moe_percentage < 5 ~ "High Confidence",
           moe_percentage >= 5 & moe_percentage <= 10 ~ "Moderate",
           moe_percentage > 10 ~ "Low Confidence"
         ))

# Create a summary showing count of counties in each reliability category
# Hint: use count() and mutate() to add percentages
count(ny_reliability, reliability)

  # why mutate?
```

## 2.3 High Uncertainty Counties

**Your Task:** Identify the 5 counties with the highest MOE percentages.

**Requirements:** - Sort by MOE percentage (highest first) - Select the
top 5 counties - Display: county name, median income, margin of error,
MOE percentage, reliability category - Format as a professional table
using `kable()`

**Hint:** Use `arrange()`, `slice()`, and `select()` functions.

```{r high-uncertainty}
# Create table of top 5 counties by MOE percentage
high_uncertainty <- ny_reliability %>% 
  filter(moe_percentage > 8) %>% 
  arrange(desc(moe_percentage)) %>% 
  slice(1:5) %>%
  select(county_name, medHHincE, medHHincM, moe_percentage, reliability)

head(high_uncertainty)

# Format as table with kable() - include appropriate column names and caption
kable(high_uncertainty,
      col.names = c("County","Median Income", "Median Income MOE", "MOE %",
                    "Reliability"),
      caption = "Counties with Highest Income Data Uncertainty",
      format.args = list(big.mark = ','))

# looking at range of values of MOE percentages
ggplot(high_uncertainty, aes(y = moe_percentage)) +
  geom_boxplot()

summary(high_uncertainty$moe_percentage)
```

**Data Quality Commentary:**

\[Write 2-3 sentences explaining what these results mean for algorithmic
decision-making. Consider: Which counties might be poorly served by
algorithms that rely on this income data? What factors might contribute
to higher uncertainty?\]

\[Counties poorly served by algorithms that rely on this income data
could be those that have a smaller population size. In these cases, the
income data could be susceptible to skewing by one or two households who
make a larger income than most other households. With greater
uncertainty, government allocation of resources, in turn, may be less
likely to occur or be more or less plentiful than for other counties in
the state, generating inefficient funding allocations.\]

# Part 3: Neighborhood-Level Analysis

## 3.1 Focus Area Selection

**Your Task:** Select 2-3 counties from your reliability analysis for
detailed tract-level study.

**Strategy:** Choose counties that represent different reliability
levels (e.g., 1 high confidence, 1 moderate, 1 low confidence) to
compare how data quality varies.

```{r select-counties}
# Use filter() to select 2-3 counties from your county_reliability data
# Store the selected counties in a variable called selected_counties
selected_counties <- ny_reliability %>%
  filter(county_name %in% c("Queens", "Schuyler", "Hamilton")) %>%
  select(county_name, medHHincE, moe_percentage, reliability)

# Display the selected counties with their key characteristics
# Show: county name, median income, MOE percentage, reliability category
head(selected_counties)
```

**Comment on the output:** \[The highest household income is associated
with the county with the highest reliability score, whereas the
remaining two 'moderate' and 'low confidence' counties show median
household incomes in the \$60,000's\]

## 3.2 Tract-Level Demographics

**Your Task:** Get demographic data for census tracts in your selected
counties.

**Requirements:** - Geography: tract level - Variables: white alone
(B03002_003), Black/African American (B03002_004), Hispanic/Latino
(B03002_012), total population (B03002_001) - Use the same state and
year as before - Output format: wide - **Challenge:** You'll need county
codes, not names. Look at the GEOID patterns in your county data for
hints.

```{r tract-demographics}
# Define your race/ethnicity variables with descriptive names
# Use get_acs() to retrieve tract-level data
# Hint: You may need to specify county codes in the county parameter
selected_demographics <- get_acs(
  geography = "tract",
  variables = c(white_alone = "B03002_003", black = "B03002_004", hisp_lat = "B03002_012", tot_pop = "B03002_001"),
  year = 2022,
  survey = "acs5",
  output = "wide",
  state = my_state,
  county = c(047, 097, 081))

# Calculate percentage of each group using mutate()
# Create percentages for white, Black, and Hispanic populations
demographics_with_pcts <- selected_demographics %>%
  mutate(pct_white = round((white_aloneE/tot_popE)*100,2),
         pct_black = round((blackE/tot_popE)*100,2),
         pct_hisplat = round((hisp_latE/tot_popE)*100,2)
  )

# Add readable tract and county name columns using str_extract() or similar
demographics_clean <- demographics_with_pcts %>%
  mutate(
    #remove Census Tract and number from county name
    county_name = str_remove(NAME, "Census Tract \\d+;"),
    county_name = str_remove(county_name, "Census Tract \\d+\\.\\d*;"),
    #remove state name and "County" word from county names
    county_name = str_remove(county_name, "County; New York"),
    #remove county and state
    tract = str_remove(NAME, ";.*"),
    tract = str_remove(tract, "Census Tract")
  )
head(demographics_clean)
```

## 3.3 Demographic Analysis

**Your Task:** Analyze the demographic patterns in your selected areas.

```{r demographic-analysis}
# Find the tract with the highest percentage of Hispanic/Latino residents
# Hint: use arrange() and slice() to get the top tract
highest_pct_hisplat <- demographics_clean %>% 
  arrange(desc(pct_hisplat)) %>% 
  slice(1)

# Calculate average demographics by county using group_by() and summarize()
# Show: number of tracts, average percentage for each racial/ethnic group

count(demographics_clean, county_name) #find total census tracts for each county


county_avg_demographics <- demographics_clean %>%
  group_by(county_name) %>%
  summarize(
    n_tracts = n(),
    avg_pct_white = round(mean(pct_white, na.rm = TRUE), 2),
    avg_pct_black = round(mean(pct_black, na.rm = TRUE), 2),
    avg_pct_hisplat = round(mean(pct_hisplat, na.rm = TRUE), 2),
    avg_tot_pop = round(mean(tot_popE, na.rm = TRUE), 2)
  )

# Create a nicely formatted table of your results using kable()
kable(county_avg_demographics,
      col.names = c("County","Total Tracts", "% White", "% Black",
                    "% Hispanic/Latino", "Total Population"),
      caption = "Queens, Schuyler and Hamilton Counties: Average Demographics",
      format.args = list(big.mark = ','))
```

# Part 4: Comprehensive Data Quality Evaluation

## 4.1 MOE Analysis for Demographic Variables

**Your Task:** Examine margins of error for demographic variables to see
if some communities have less reliable data.

**Requirements:** - Calculate MOE percentages for each demographic
variable - Flag tracts where any demographic variable has MOE \> 15% -
Create summary statistics

```{r demographic-moe}
# Calculate MOE percentages for white, Black, and Hispanic variables
# Hint: use the same formula as before (margin/estimate * 100)

# Create a flag for tracts with high MOE on any demographic variable
# Use logical operators (| for OR) in an ifelse() statement

county_dems_reliability <- demographics_clean %>%
  # using ifelse() to guard against zeros that will give Inf results, messing up plots
  mutate(moe_pct_white = ifelse(pct_white > 0, round((white_aloneM / white_aloneE)*100, 2), NA), 
         moe_pct_black = ifelse(pct_black > 0, round((blackM / blackE)*100, 2), NA),
         moe_pct_hsp = ifelse(pct_hisplat > 0, round((hisp_latM / hisp_latE)*100, 2), NA),
         high_moe_flag = ifelse(
             moe_pct_white > 15 |
             moe_pct_black > 15 |
             moe_pct_hsp > 15,
             1, 0
             )
  )
# assessing variability of each demographic variable moe percentage values
summary(county_dems_reliability$moe_pct_white) 
summary(county_dems_reliability$moe_pct_black)
summary(county_dems_reliability$moe_pct_hsp)

# Create summary statistics showing how many tracts have data quality issues
tracts_reliability_summary <- county_dems_reliability %>%
  group_by(county_name) %>%
  summarize(
    total_tracts = n(),
    tracts_with_issues = sum(high_moe_flag, na.rm = TRUE),
    pct_tracts_with_issues = mean(high_moe_flag, na.rm = TRUE) * 100
  )
head(tracts_reliability_summary)
  # total_tracts and tracts_with_issues are not equal but the pct_tracts_with_issues are all 100% - the "missing" tracts are those with "NA"      values

```

## 4.2 Pattern Analysis

**Your Task:** Investigate whether data quality problems are randomly
distributed or concentrated in certain types of communities.

```{r pattern-analysis}
# Group tracts by whether they have high MOE issues
tract_dems_reliability <- county_dems_reliability %>%
  mutate(
    white_reliability = case_when(
      moe_pct_white <= 20 ~ "High",
      moe_pct_white <= 40 ~ "Moderate",
      moe_pct_white <= 60 ~ "Low",
      TRUE ~ "Very low" # if no above conditions are met, assign tract "Very low" category; without this case_when() would return NAs
    ),
    black_reliability = case_when(
      moe_pct_black <= 20 ~ "High",
      moe_pct_black <= 40 ~ "Moderate",
      moe_pct_black <= 60 ~ "Low",
      TRUE ~ "Very low"
    ),
    hispanic_reliability = case_when(
      moe_pct_hsp <= 20 ~ "High",
      moe_pct_hsp <= 40 ~ "Moderate",
      moe_pct_hsp <= 60 ~ "Low",
      TRUE ~ "Very low"
    )
  )

high_uncertainty_tracts <- tract_dems_reliability %>%
  filter(white_reliability == "Low", 
         black_reliability == "Low", 
         hispanic_reliability == "Low") %>% 
  slice(1:5) %>%
  select(tract, pct_white, pct_black, pct_hisplat, tot_popE, white_reliability)

view(high_uncertainty_tracts)

# Calculate average characteristics for each group - population size, demographic percentages
# Use group_by() and summarize() to create this comparison
  # white population - analysis
white_reliability_summary <- tract_dems_reliability %>%
  group_by(white_reliability) %>%
  summarize(
    n_tracts = n(),
    avg_pct_white = round(mean(pct_white), 2),
    avg_tot_pop = round(mean(tot_popE), 2)
  )
view(white_reliability_summary)
  # black population - analysis
black_reliability_summary <- tract_dems_reliability %>%
  group_by(black_reliability) %>%
  summarize(
    n_tracts = n(),
    avg_pct_black = round(mean(pct_black), 2),
    avg_tot_pop = round(mean(tot_popE), 2)
  )
view(black_reliability_summary)
  # hispanic/latino population - analysis
hsplat_reliability_summary <- tract_dems_reliability %>%
  group_by(hispanic_reliability) %>%
  summarize(
    n_tracts = n(),
    avg_pct_hsplat = round(mean(pct_hisplat), 2),
    avg_tot_pop = round(mean(tot_popE), 2)
    )
view(hsplat_reliability_summary)

# Create a professional table(s) showing the patterns
kable(white_reliability_summary,
      col.names = c("Reliability","Total Tracts", "% White", "Total Population"),
      caption = "Tract Average Demographic Characteristics by Data Reliability Levels",
      format.args = list(big.mark = ','))
kable(black_reliability_summary,
      col.names = c("Reliability","Total Tracts", "% Black", "Total Population"),
      caption = "Tract Average Demographic Characteristics by Data Reliability Levels",
      format.args = list(big.mark = ','))
kable(hsplat_reliability_summary,
      col.names = c("Reliability","Total Tracts", "% Hispanic/Latino", "Total Population"),
      caption = "Tract Average Demographic Characteristics by Data Reliability Levels",
      format.args = list(big.mark = ','))
kable(high_uncertainty_tracts,
      col.names = c("Tract", "% White", "% Black", "% Hispanic/Latino", "Total Population", "Reliability"),
      caption = "Top 5 Tracts with Highest Data Uncertainty: Average Demographics",
      format.args = list(big.mark = ','))
```

**Pattern Analysis:** \[The tracts that tend to fall in the "high"
reliability category are those with total populations that are on
average above 3,500. The specific demographic race/ethnicity variables
are also most reliable when the average population of the respective
demographic groups are on average greater than 2000. The demographic
estimates start to lose reliability (low and very low) when the average
population amounts are below 700. Such observations align with the logic
that error margins tend to increase, and therefore reliability
decreases, when sample populations are small.\]

# Part 5: Policy Recommendations

## 5.1 Analysis Integration and Professional Summary

**Your Task:** Write an executive summary that integrates findings from
all four analyses.

**Executive Summary Requirements:** 1. **Overall Pattern
Identification**: What are the systematic patterns across all your
analyses? 2. **Equity Assessment**: Which communities face the greatest
risk of algorithmic bias based on your findings? 3. **Root Cause
Analysis**: What underlying factors drive both data quality issues and
bias risk? 4. **Strategic Recommendations**: What should the Department
implement to address these systematic issues?

**Executive Summary:**

\[1. Across all the analyses, places with low populations tend to
produce higher margins of error for a given estimate. For example,
median income values were least reliable in sparsely populated counties.
The same was true for estimates of total populations of different race
and ethnic groups across tracts, where the lowest reliability was
associated with tracts where both a respective demographic group (white,
black, hispanic/latino) represented a small share of the populationn and
the total population of the census tract was low. 2. Based on these
findings, the communities which face the greatest risk of algorithmic
bias are those with low total populations and those where a specific
racial or ethnic group's population is also low. 3. One root cause of
data quality issues associated with low populated areas is the
difficulty of surveyors to reach their populations of interest. In more
rural, remote places, peoples' physical addresses may be inconsistent or
missing. 4. The Department might consider greater in-person contact with
small, more remote communities to develop community-informed strategies
to best reach residents in the area, perhaps through engagement with
community spaces like faith-based gathering centers.\]

## 6.3 Specific Recommendations

**Your Task:** Create a decision framework for algorithm implementation.

```{r recommendations-data}
# Create a summary table using your county reliability data
# Include: county name, median income, MOE percentage, reliability category
summary_table <- ny_reliability %>%
  select(county_name, medHHincE, moe_percentage, reliability)

# Add a new column with algorithm recommendations using case_when():
# - High Confidence: "Safe for algorithmic decisions"
# - Moderate Confidence: "Use with caution - monitor outcomes"  
# - Low Confidence: "Requires manual review or additional data"
summary_table1 <- summary_table %>%
  mutate(recommendation = case_when(
    moe_percentage < 5 ~ "Safe for algorithmic decisions",
    moe_percentage >= 5 & moe_percentage <= 10 ~ "Use with caution - monitor outcomes",
    moe_percentage > 10 ~ "Requires manual review or additional data"
  ))

summary_table1

view(summary_table1)
# Format as a professional table with kable()
kable(summary_table1,
      col.names = c("County", "Median Income", "% MOE", "Reliability", "Recommendation"),
      caption = "NY County Median Income Data: Reliability and Recommendations",
      format.args = list(big.mark = ','))

```

**Key Recommendations:**

**Your Task:** Use your analysis results to provide specific guidance to
the department.

1.  **Counties suitable for immediate algorithmic implementation:**
    \[Albany, Allegany, Bronx, Broome, and Cattaraugus are all counties
    in NY that have high confidence data, meaning that they are
    appropriate for algorithmic implementation as there will be little
    concern that any analyzed values are mis-representative of reality
    due to small MOE percentages.\]

2.  **Counties requiring additional oversight:** \[Essex, Greene,
    Schuyler, Seneca, and Yates are counties with moderate confidence
    data that would require some additional effort to verify whether the
    values provide an accurate reflection of reality in these areas,
    perhaps by supplementing the analysis of median income using proxy
    data such as home values.\]

3.  **Counties needing alternative approaches:** \[ Hamilton county has
    low confidence median income data that will definitely require
    further survey collection to boost the count of the sample
    population in order to make reasonable inferences with the resulting
    data.\]

## Questions for Further Investigation

\[1. How does one account for places where populations have recently
fluctuated rapidly but their populations are large, likely producing
smaller MOEs? 2. How have MOEs changed or not for key racial and ethnic
groups over time?\]

# Technical Notes

**Data Sources:** - U.S. Census Bureau, American Community Survey
2018-2022 5-Year Estimates - Retrieved via tidycensus R package on
\[2/5/2026\]

**Reproducibility:** - All analysis conducted in R version
\[2026.01.0+392\] - Census API key required for replication - Complete
code and documentation available at:
\[<https://github.com/maudeceruso/MUSA5080_Portfolio>\]

**Methodology Notes:** None

**Limitations:** None

------------------------------------------------------------------------

## Submission Checklist

Before submitting your portfolio link on Canvas:

-   [ ] All code chunks run without errors
-   [ ] All "\[Fill this in\]" prompts have been completed
-   [ ] Tables are properly formatted and readable
-   [ ] Executive summary addresses all four required components
-   [ ] Portfolio navigation includes this assignment
-   [ ] Census API key is properly set
-   [ ] Document renders correctly to HTML

**Remember:** Submit your portfolio URL on Canvas, not the file itself.
Your assignment should be accessible at
`your-portfolio-url/labs/lab_1/your_file_name.html`
